### 2017-09-18 paper/paper.tex
Add paper.

\documentclass[preprint,12pt]{elsarticle}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{lineno}

\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}

\begin{document}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\title{Multi-output Bus Travel Time Prediction Using Convolutional LSTM Neural Network}

\journal{Expert Systems with Applications}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{}
%% \address[label1]{}
%% \address[label2]{}

\author{Niklas Christoffer Petersen}

\address{
  Department of Management Engineering \\
  Technical University of Denmark, 2800 Kongens Lyngby \\
  niklch@dtu.dk
}

\begin{abstract}
This paper presents a multi-output, multi-time-step, deep neural network for bus travel time prediction using Convolutional and Long short-term memory (LSTM) layers. The method is evaluated and compared to other popular approaches for link travel time prediction.
\end{abstract}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
Bus Travel Time Prediction \sep Intelligent Transport Systems \sep Convolutional Neural Network (CNN) \sep Long short-term memory (LSTM)
\end{keyword}

\end{frontmatter}
\linenumbers
\section{Introduction}
Public transport authorities has long found that GPS trajectory data from already deployed \emph{Automatic Vehicle Location}-systems (AVL) can be used in \emph{Intelligent Transport Systems} (ITS) cf.\ \cite{Tcrp48}. Examples include real-time traffic information for passengers, including arrival- and departure time predictions, where studies \cite{Fan2016} show that reliable real-time information at bus stops has a statistical significant dampening effect on the perceived waiting time.

Besides real-time passenger information, robust arrival- and departure time predictions are necessary for operating more sophisticated ITS applications successfully, e.g.\ demand-adaptive transit systems, and \emph{connection insurance} between public transport services. Arrival/departure time prediction is commonly approached as a specialization of travel time prediction, where the predicted travel time is simply accumulated downstream the route to yield the arrival/departure time predictions at each stop point of the rest of the current journey. Besides the link travel time, estimations of dwell time (i.e.\ when a bus is holding at a stop point) are accumulated downstream.

Producing precise bus travel time predictions in areas with little external influence, e.g.\ rural areas, can be solved to a large extent with historical averaging or simple regression methods \cite{Williams2003,Altinkaya2013}. The problem becomes much more complex in urban areas where congestion, events, road-works, weather, etc.\ highly influences the traffic flow and passenger demand.

This paper presents a multi-output, multi-time-step deep neural network for bus travel time prediction using \emph{Convolutional} and \emph{Long short-term memory} (LSTM) \cite{Lstm1,Lstm2} layers. The goal of this work is to produce precise short-term predictions (e.g.\ 0--3 hours) for link travel time, specifically for bus traffic in urban areas. The method is evaluated and compared to other popular approaches for link travel time prediction. The paper is structured in the following manner: In the next related work and literature is reviewed. \Cref{sec:convlstm} introduces Convolutional LSTM neural networks in general and in \Cref{sec:model} we present the proposed multi-output model in more details, including e.g.\ network topology, etc. \Cref{sec:experiments} introduces the dataset which the model has been evaluated on and results are presented and discussed in~\Cref{sec:results}. Finally we conclude on the work in~\Cref{sec:conclusion}.

\subsection{Literature review}
Early approaches presents historical averaging models \cite{Dailey1999,Sun2007}, and linear regression \cite{Patnaik2004}. Recent research presents this type of model only for comparison purpose, and the models are in all cases outperformed by the compared models \cite{Shalaby2004,Jeong2005}. Kalman-filters, which by its capabilities of maintaining state between predictions, has shown interest from several studies, either as an independent model \cite{Chen2001,Shalaby2004} or in combination with other models \cite{Zaki2013,Bai2015}.

Both the different regression style models and the Kalman-filter models still has limited options for capturing fluctuations in travel and dwell time in a metropolitan bus system, since they to a large extent still is averaging and smoothing their response. E.g., the Kalman-filter's state is only directly accessible for the leading time step, and thus not capable of finding long-distance patterns spanning over several links and/or over several time steps. This is substantiated by \cite{Lin2013} and \cite{Kumar2014} that finds artificial neural networks (ANN) outperforms Kalman-filter models.

The computational challenges of pure ANN approaches i.a.\ has sparked the interest for studying composite or hybrid models. \cite{Zaki2013} uses a two-stage approach by combining offline ANN-models, with an adaptable/online Kalman-filter to yield a dynamic model. The model is not actually learning in the long term, but is able to adapt to temporal variations in the current travel time on a journey.

Some recent research recognizes that several routes can benefit from each other's predictions if they share some partial route segment, e.g.\ \cite{Yu2011,Gal2014,Bai2015}, however none of these approaches consider cross temporal correlations between different route segments, and only uses a small window for correlation with upstream links (e.g.\ max. 3 links).

\cite{YanjieDuan2016} proposes to use LSTM model for general highway travel time prediction, and do predict multiple steps ahead, but only for a single link at a time, i.e.\ cross link (spacial) correlations are lost.

\section{Convolutional LSTM neural networks}
\label{sec:convlstm}
A Long short-term memory (LSTM) neural network is a special type of Recurrent Neural Network~(RNN) which has proven robust for capturing long-distance dependencies \cite{Lstm1,Lstm2}. The important feature of a LSTM network is its capability to persist cell state, $c_t$, from previous observations across sequences of input (e.g.\ time), but also eliminate information which is considered unimportant. To allow this mechanism the persistence of information is controlled by three gates: \emph{input gate}, \emph{forget gate}, and \emph{output gate}. Each gate yields a state value at time $t$, respectively $i_t$, $f_t$, and $o_t$, along with the cell output, $h_t$, cf.\ \Cref{eq:lstm}.

\begin{equation}
\begin{aligned}
  i_t &= \sigma \left( W_\mathit{xi} x_t + W_\mathit{hi} h_{t-1} + W_{ci} c_{t-1} + b_i \right) \\
  f_t &= \sigma \left( W_\mathit{xf} x_t + W_\mathit{hf} h_{t-1} + W_\mathit{cf} c_{t-1} + b_f \right) \\
  c_t &= f_t c_{t-1} + i_t \mathrm{tanh} \left( W_\mathit{xc} x_t + W_\mathit{hc} h_{t-1} + b_c \right) \\
  o_t &= \sigma \left( W_\mathit{xo} x_t + W_\mathit{ho} h_{t-1} + W_\mathit{co} c_{t} + b_o \right) \\
  h_t &= o_t \mathrm{tanh} \left( c_t \right)
\end{aligned}
\label{eq:lstm}
\end{equation}

\Cref{fig:lstm-peepholes} illustrates the inner structure of a LSTM cell with peephole as proposed by \cite{LstmPeephole}. 
Is has especially grown popular for predicting time series using methods evolved from \cite{LstmTs}, where fixed-length windows of time-series are generated and feed into a LSTM network. Multiple LSTMs can be stacked such more complex patterns of sequential information (e.g.\ temporal patterns) can be learned.

\begin{figure}[!ht]
  \centering
  \includegraphics[scale=.6]{drawings/lstm-peepholes.pdf}
  \caption{Structure of LSTM cell with peephole.}
  \label{fig:lstm-peepholes}
\end{figure}

Convolutional Neural Networks~(CNN) on the other hand has been widely used for capturing spacial relationships, e.g.\ importance of neighboring pixels in an image. As opposed to fully connected layers, where each unit, $i$, in the layer, has a dedicated weight, $w_{ij}$, for all input, $x_j$, convolutional units are only locally connected and reuses the same weights for producing several outputs. Instead of considering the entire input-vector, only a fixed sized window, or \emph{convolution}, around each input is considered. The weights are therefore referred to as the \emph{filters} or \emph{kernels} of the layer. \Cref{fig:conv} illustrates a single convolutional filter of size $3$ being applied to 1-dimensional data.

\begin{figure}[!ht]
  \centering
  \includegraphics[scale=.6]{drawings/conv.pdf}
  \caption{Application of Convolutional filter onto 1D data.}
  \label{fig:conv}
\end{figure}

Special care needs to be taken on the boundaries, i.e.\ where the convolutional filter will exceed the input. A popular approach to avoid the size of the output decreases is to pad the input, e.g.\ with zeros. This ensures that the output shape of each convolutional unit will always be identical to the input shape, which is often desirable. One of the key benefits of convolutional networks is that the number of weights that needs to be learned is greatly reduced compared to fully connected networks, and that learned patterns transfers across space. I.e.\ the convolutional filters become feature detectors, that in our case can detect spacial patterns across links, e.g.\ congestion forming, etc.

\cite{ConvLSTM} introduced the novel combination of Convolutional and LSTM layers into a single structure, the \emph{Convolutional LSTM}, or simply \emph{ConvLSTM}. Specifically the method applies convolutional filters in the input-to-state and state-to-state transitions the LSTM cf.\ \Cref{eq:convlstm}, where $*$ denotes the convolution operator.

\begin{equation}
\begin{aligned}
  i_t &= \sigma \left( W_\mathit{xi} * x_t + W_\mathit{hi} * h_{t-1} + W_{ci} c_{t-1} + b_i \right) \\
  f_t &= \sigma \left( W_\mathit{xf} * x_t + W_\mathit{hf} * h_{t-1} + W_\mathit{cf} c_{t-1} + b_f \right) \\
  c_t &= f_t c_{t-1} + i_t \mathrm{tanh} \left( W_\mathit{xc} * x_t + W_\mathit{hc} * h_{t-1} + b_c \right) \\
  o_t &= \sigma \left( W_\mathit{xo} * x_t + W_\mathit{ho} * h_{t-1} + W_\mathit{co} c_{t} + b_o \right) \\
  h_t &= o_t \mathrm{tanh} \left( c_t \right)
\end{aligned}
\label{eq:convlstm}
\end{equation}

\section{Multi-output model}
\label{sec:model}

Input data is arranged in $N$ samples, each with $w$ time steps $t-w+1, \ldots, t$, and each time step with $k$ link travel times $e_1, \ldots, e_k$ cf.~\Cref{fig:data_shape}. Likewise is the output arranged with predictions of $j$ time steps ahead, $t+1, \ldots, t+j$. Thus the input is a 4D-tensor with shape $(N, w, k, 1)$, and the output a 4D-tensor with shape $(N, j, k, 1)$. It is emphasized that each prediction consists of travel time predictions for all links for the next $j$ time steps, i.e.\ multi-output, multi-time-step prediction.
\begin{figure}[!ht]
  \centering
  \includegraphics[width=0.6\textwidth]{drawings/data_shape-crop.pdf}
  \caption{Shapes of the input and output data.}
  \label{fig:data_shape}
\end{figure}

Obviously the travel time varies throughout the time of the day, and the day of the week due to \emph{recurring congestion}. In order to reduce the need for the network to learn this recurring variation, data is normalized to focus on deviations from the normal and expected pattern. Travel times are centered with the mean for each link, at teh time of day, and day of week, $\mathit{\bar{x}_{l,\mathit{dow},\mathit{tod}}}$, and scaled with the standard for each link, $\sigma_l$, cf.\ \Cref{eq:normalization}.

\begin{equation}
  \frac{x - \bar{x}_{l,\mathit{dow},\mathit{tod}}}{\sigma_l}
  \label{eq:normalization}
\end{equation}

\subsection{Network topology and training}
\Cref{fig:ConvLSTM} shows the overall network topology, where blue boxes illustrates \emph{input-to-state} convolutions and yellow \emph{state-to-state} convolutions. The network uses an encoder/decoder technique that to a large extent follows \cite{ConvLSTM}, where the encoder-block consists of two \emph{ConvLSTM} layers. The result is feed into a decoder, or prediction block, also consisting of two \emph{ConvLSTM} layers. The architecture allows unequal $w$ and $k$, e.g.\ predict the next $3$ time steps based on a window size of $10$.

Thus convolutional filters are applied to each input, at each time step to the respective LSTM-cell, and also between LSTM-cells in the state-transition. Since the time steps are one-dimensional (link travel times across links), the filters are also one-dimensional. In total the 4 layers of Convolutional LSTMs are arranged with filter sizes of $5\times1$ for both \emph{input-to-state} and \emph{state-to-state} filters.

To avoid over-fitting doing training \emph{Dropout} \cite{Dropout} is performed between the \emph{ConvLSTM} layers, and \emph{Batch Normalization} \cite{BatchNorm} are also performed before each \emph{ConvLSTM} layers to ensure reasonable input for activations. The dropout factor is adjusted to respectively 20\%, 10\% and 10\%.

\begin{figure}[!ht]
  \centering
  \includegraphics[scale=.8]{drawings/conv_lstm.pdf}
  \caption{Convolutional LSTM network topology.}
  \label{fig:ConvLSTM}
\end{figure}

The network is trained using the \emph{RMSprop}-algorithm \cite{RMSprop}.

\section{Experiments}
\label{sec:experiments}
For evaluation the method is applied to a dataset from Copenhagen's public transport authority. The dataset consists of 814,710~travel time observations for the ``4A'' bus line in the period January 1 to May 31 2017. The data points were  collected using the real-time AVL-system installed in all vehicles of the line. The geography of the route is shown in \Cref{fig:4a_map}. As the line circles central Copenhagen, it is potentially highly sensitive to congestion to/from the city as it intersects with several large corridors along its route.

TODO: How much used for training, how much for testing?

\begin{figure}[!ht]
  \centering
  \includegraphics[width=0.6\textwidth]{drawings/4a.png}
  \caption{Geography of the 4A bus line in Copenhagen.}
  \label{fig:4a_map}
\end{figure}

\subsection{Data preprocessing}
Since the travel time measurement contains several extreme outliers due to measurement faults, these were labeled and dropped using the \emph{absolute deviation around the median} (MAD) cf.\ \cite{Olewuezi2011}. Data is aggregated into 15-minute time steps and normalized cf.~\Cref{sec:model}.

The implementation of the proposed model was done in Python using Keras \cite{Keras}.

\subsection{Evaluation}
In order to evaluate the presented model and comparisons, the following measures is used: \emph{mean absolute error} (MAE), \emph{root mean square error} (RMSE), and \emph{mean absolute percentage error} (MAPE) cf.~\Cref{eq:mae,eq:rmse,eq:mape}, where $X_i$ is the true travel time for sample $i$ and $\widehat{X}_i$ is the predicted travel time.

\begin{equation}
    \textrm{MAE}(X, \widehat{X}) = \frac{\sum_{i = 1}^{n} \left| X_i - \widehat{X}_i \right| }{n}
    \label{eq:mae}
\end{equation}

\begin{equation}
    \textrm{RMSE}(X, \widehat{X}) = \sqrt{\frac{\sum_{i = 1}^{n} \left(X_i - \widehat{X}_i \right)^2}{n}}
    \label{eq:rmse}
\end{equation}

\begin{equation}
    \textrm{MAPE}(X, \widehat{X}) = \frac{1}{n} \sum_{i = 1}^{n} \left| \frac{X_i - \widehat{X}_i}{X_i} \right| 
    \label{eq:mape} 
\end{equation}

\section{Results and discussion}
\label{sec:results}
The model is trained on the prepared data using a sliding window approach to simulate real-world conditions where real-time travel time measurements arrives as a continuously data stream. \Cref{fig:loss} shows the evolution of training and validation loss for each epoch for such a window, where the loss is the MAE if the individual links.

\begin{figure}
  \centering
  TODO: Needs to be updated ...
  \includegraphics[width=0.8\textwidth]{plots/conv_lstm_model_loss.pdf}
  \caption{Training and validation loss evolution.}
  \label{fig:loss}
\end{figure}

\Cref{tab:results} shows the results of the presented method and compares to two other methods: 1) a naïve historical average model, i.e. equivalent of just prediction the normalized value, $\bar{x}_{l,\mathit{dow},\mathit{tod}}$ and 2) a pure LSTM model as proposed by \cite{YanjieDuan2016}. Predictions are accumulated downstream on journey level to simulate for use in real-time bus arrival/departure time prediction.

From the table it is found that the Convolutional LSTM performs better than the compared methods. Although the advantage might seem small it is emphasized that evaluation measurements are averaging their response, and thus the increased accuracy can be much higher on individual journeys, especially if they experienced very irregular travel times.

\begin{table}[!ht]
    \caption{Results of the presented and other popular models}
    \label{tab:results}
    \center
    \begin{tabular}{ll|rrr}
        Model & Time ahead & MAE (min) & RMSE (min) & MAPE (\%) \\
        \hline 
        Historical average &                & 4.90 & 6.32 & 7.38 \% \\ 
        \hline 
        Pure LSTM          & t + 1 (15 min) & 3.71 & 3.71 & 5.87 \% \\
        Pure LSTM          & t + 2 (30 min) & 4.31 & 5.63 & 6.63 \% \\
        Pure LSTM          & t + 3 (45 min) & 4.97 & 6.41 & 7.52 \% \\
        \hline 
        Convolutional LSTM & t + 1 (15 min) & 2.96 & 3.94 & 5.10 \% \\
        Convolutional LSTM & t + 2 (30 min) & 3.04 & 4.08 & 5.21 \% \\
        Convolutional LSTM & t + 3 (45 min) & 3.17 & 4.24 & 5.37 \% \\
        \hline 
    \end{tabular}
\end{table}

\subsection{Future work}
For the prediction accuracy to be increased further it is proposed to apply ensemble/multi-model approaches. In this case the presented model can be included and used as a sub-model for the ensemble. Further research should be invested in the more rare, but highly impacting deviations, e.g.\ traffic incidents, extreme weather conditions, etc.

\section{Conclusion}
\label{sec:conclusion}
This paper has presented a multi-output, multi-time-step deep neural network for bus travel time prediction using Convolutional and Long short-term memory (LSTM) layers. Results shown the presented network outperforms other popular and recent methods.

Future research opportunities include using the model a part of a multi-model ensemble, and focus in the more rare, but highly impacting deviations.

\newpage

\bibliographystyle{elsarticle-num}
%\bibliography{../references/library}
\bibliography{c:/development/references/library}
\end{document}

### 2017-09-26 paper/paper.tex
Add figures.

@@
 \section{Introduction}
-Public transport authorities has long found that GPS trajectory data from already deployed \emph{Automatic Vehicle Location}-systems (AVL) can be used in \emph{Intelligent Transport Systems} (ITS) cf.\ \cite{Tcrp48}.
-Examples include real-time traffic information for passengers, including arrival- and departure time predictions, where studies \cite{Fan2016} show that reliable real-time information at bus stops has a statistical significant dampening effect on the perceived waiting time.
+Public transport authorities has long found that GPS trajectory data from already deployed \emph{Automatic Vehicle Location}-systems (AVL) can be used in \emph{Intelligent Transport Systems} (ITS) \cite{Tcrp48}.
+Examples include real-time traffic information for passengers, e.g.\ departure boards, where studies has shown that reliable real-time information at bus stops has a statistical significant dampening effect on the perceived waiting time \cite{Fan2016}.
 
-Besides real-time passenger information, robust arrival- and departure time predictions are necessary for operating more sophisticated ITS applications successfully, e.g.\ demand-adaptive transit systems, and \emph{connection insurance} between public transport services.
+Besides real-time passenger information, robust arrival- and departure time predictions are necessary for operating more sophisticated ITS applications successfully, e.g.\ demand-adaptive transit systems, and \emph{connection insurance} between different public transport services.
 Arrival/departure time prediction is commonly approached as a specialization of travel time prediction, where the predicted travel time is simply accumulated downstream the route to yield the arrival/departure time predictions at each stop point of the rest of the current journey.
@@
 
-This paper presents a multi-output, multi-time-step deep neural network for bus travel time prediction using \emph{Convolutional} and \emph{Long short-term memory} (LSTM) \cite{Lstm1,Lstm2} layers.
+In this paper we presents a multi-output, multi-time-step deep neural network for bus travel time prediction using \emph{Convolutional} and \emph{Long short-term memory} (LSTM) \cite{Lstm1,Lstm2} layers.
 The goal of this work is to produce precise short-term predictions (e.g.\ 0--3 hours) for link travel time, specifically for bus traffic in urban areas.
@@
   \centering
-  \includegraphics[width=0.6\textwidth]{drawings/data_shape-crop.pdf}
+  \includegraphics[width=0.8\textwidth]{drawings/data_shape-crop.pdf}
   \caption{Shapes of the input and output data.}

### 2017-09-26 jupyter/common.py
Updating common functions.

@@ -1,0 +1,45 @@
+import numpy as np
+import pandas as pd
+
+from scipy.stats import rankdata
+
+def prep_data(file, sep = ';', ref_freq = '15min'):
+    data = pd.read_csv(file, sep)
+
+    data.sort_values('DateTime', inplace = True)
+
+    # Add a reference date for grouping similar measurements
+    ix = pd.DatetimeIndex(pd.to_datetime(data['DateTime'])).floor(ref_freq)
+    
+    data["DateTimeRef"] = ix
+    data["DowTimeRef"] = rankdata(data['DayOfWeek'].values * 10000 + ix.hour*100 + ix.minute, 'dense')
+    
+    return data
+
+def fit_scale(data, smooth = 7):
+    means = { }
+    scales = { }
+    low = { }
+    upr = { }
+
+    for k, v in data[~np.isnan(data['LinkTravelTime'])].sort_values('LineDirectionLinkOrder').groupby('LinkRef', sort = False):
+        median = np.median(v['LinkTravelTime']);
+        mad = 1.4826 * np.median(np.abs(v['LinkTravelTime'] - median))
+        low[k] = max(median - 3 * mad, 0)
+        upr[k] = median + 3 * mad
+
+        mean = v[(low[k] < v['LinkTravelTime']) & (v['LinkTravelTime'] < upr[k])].groupby("DowTimeRef")["LinkTravelTime"].mean()
+        mean = mean.interpolate().rolling(window = smooth, center = True).mean()
+        means[k] = mean
+        #scales[k] = v_[(low[k] < v_['LinkTravelTime']) & (v_['LinkTravelTime'] < upr[k])]['LinkTravelTime'].std()
+        scales[k] = 1
+        
+    means_df = pd.DataFrame(data = means).fillna(method='pad').fillna(method='bfill')
+    return (means_df, scales, low, upr)
+
+def remove_outliers(data, low, upr): 
+    _low = data['LinkRef'].apply(lambda k: low[k])
+    _upr = data['LinkRef'].apply(lambda k: upr[k])
+    mask = ~(_low < data['LinkTravelTime']) & (data['LinkTravelTime'] < _upr)
+    data.loc[mask, 'LinkTravelTime'] = np.nan
+    return mask.sum()

### 2017-11-02 jupyter/visualization.py
Wörk, wörk, wörk.

@@ -1,0 +1,37 @@
+import matplotlib.pyplot as plt
+import matplotlib.font_manager as font_manager
+from cycler import cycler
+
+colors = {
+    'Bl':(0,0,0),
+    'Or':(.9,.6,0),
+    'SB':(.35,.7,.9),
+    'bG':(0,.6,.5),
+    'Ye':(.95,.9,.25),
+    'Bu':(0,.45,.7),
+    'Ve':(.8,.4,0),
+    'rP':(.8,.6,.7),
+}
+
+def set_plot_style():
+    plt.style.use('ggplot')
+    plt.rcParams['font.size'] = 14
+    plt.rcParams['figure.figsize'] = (16, 10)
+
+    plt.rcParams['axes.facecolor'] = 'w'
+    plt.rcParams['axes.labelcolor'] = 'k'
+    plt.rcParams['axes.edgecolor'] = 'k'
+    plt.rcParams['axes.prop_cycle'] = cycler('color',  [colors[k] for k in ['Bl', 'Or', 'SB', 'bG', 'Ye', 'Bu', 'Ve', 'rP']])
+
+    plt.rcParams['grid.color'] = (.7, .7, .7, 0)
+
+    plt.rcParams['xtick.color'] = 'k'
+    plt.rcParams['xtick.labelsize'] = 16
+    plt.rcParams['ytick.color'] = 'k'
+    plt.rcParams['ytick.labelsize'] = 16
+
+    plt.rcParams['legend.fontsize'] = 16
+    plt.rcParams['legend.frameon'] = True
+    plt.rcParams['legend.framealpha'] = 1
+    plt.rcParams['legend.facecolor'] = 'w'
+    plt.rcParams['legend.edgecolor'] = (.7, .7, .7, 0)

### 2017-11-02 paper/paper.tex
Wörk, wörk, wörk.

@@
-\documentclass[preprint,12pt]{elsarticle}
+\documentclass[preprint,11pt,5p,twocolumn]{elsarticle}
 \usepackage{amssymb}
@@
 \usepackage{graphicx}
+\usepackage{caption}
+\usepackage{subcaption}
 \usepackage{lineno}
@@
 
+\usepackage{afterpage}
+
+\makeatletter
+\providecommand{\doi}[1]{%
+  \begingroup
+    \let\bibinfo\@secondoftwo
+    \urlstyle{rm}%
+    \href{http://dx.doi.org/#1}{%
+      doi:\discretionary{}{}{}%
+      \nolinkurl{#1}%
+    }%
+  \endgroup
+}
+\makeatother
+
 \begin{document}
+
+\hyphenation{Conv-LSTM}
+\hyphenation{Conv-LSTMs}
 
@@
 
-\title{Multi-output Bus Travel Time Prediction Using Convolutional LSTM Neural Network}
+\title{Multi-output Bus Travel Time Prediction\\
+Using Convolutional LSTM Neural Network}
 
@@
 \begin{abstract}
+Accurate and reliable predictions for travel times in a public transport networks is essential for delivering an attractive service.
 This paper presents a multi-output, multi-time-step, deep neural network for bus travel time prediction using Convolutional and Long short-term memory (LSTM) layers.
-The method is evaluated and compared to other popular approaches for link travel time prediction.
+The method is evaluated and compared to other popular approaches for link travel time prediction, and currently available services.
 \end{abstract}
@@
 \end{frontmatter}
-\linenumbers
 \section{Introduction}
@@
 Arrival/departure time prediction is commonly approached as a specialization of travel time prediction, where the predicted travel time is simply accumulated downstream the route to yield the arrival/departure time predictions at each stop point of the rest of the current journey.
-Besides the link travel time, estimations of dwell time (i.e.\ when a bus is holding at a stop point) are accumulated downstream.
+Besides the link travel time, estimations of dwell time (i.e.\ when a bus is holding at a stop point) are also accumulated downstream.
 
@@
 
-In this paper we presents a multi-output, multi-time-step deep neural network for bus travel time prediction using \emph{Convolutional} and \emph{Long short-term memory} (LSTM) \cite{Lstm1,Lstm2} layers.
-The goal of this work is to produce precise short-term predictions (e.g.\ 0--3 hours) for link travel time, specifically for bus traffic in urban areas.
-The method is evaluated and compared to other popular approaches for link travel time prediction.
-The paper is structured in the following manner: In the next related work and literature is reviewed.
+In this paper we presents a multi-output, multi-time-step deep neural network for bus travel time prediction using a combination of \emph{Convolutional} and \emph{Long short-term memory} (LSTM) \cite{Lstm1,Lstm2} layers.
+The goal of this work is to produce precise short-term predictions (e.g.\ $0-1.5$ hours) for link travel time, specifically for bus traffic in urban areas.
+The method is evaluated and compared to other popular approaches for link travel time prediction, and currently available services.
+
+The paper is structured in the following manner: In the next section related work and literature is reviewed.
 \Cref{sec:convlstm} introduces Convolutional LSTM neural networks in general and in \Cref{sec:model} we present the proposed multi-output model in more details, including e.g.\ network topology, etc.
-\Cref{sec:experiments} introduces the dataset which the model has been evaluated on and results are presented and discussed in~\Cref{sec:results}.
+\Cref{sec:experiments} introduces the Copenhagen-dataset, which the model has been evaluated on, and our results are presented and discussed in~\Cref{sec:results}.
 Finally we conclude on the work in~\Cref{sec:conclusion}.
@@
 Both the different regression style models and the Kalman-filter models still has limited options for capturing fluctuations in travel and dwell time in a metropolitan bus system, since they to a large extent still is averaging and smoothing their response.
-E.g., the Kalman-filter's state is only directly accessible for the leading time step, and thus not capable of finding long-distance patterns spanning over several links and/or over several time steps.
+E.g., the Kalman-filter's state is only directly accessible for the leading time-step, and thus not capable of finding long-distance patterns spanning over several links and/or over several time-steps.
 This is substantiated by \cite{Lin2013} and \cite{Kumar2014} that finds artificial neural networks (ANN) outperforms Kalman-filter models.
@@
 The computational challenges of pure ANN approaches i.a.\ has sparked the interest for studying composite or hybrid models.
-\cite{Zaki2013} uses a two-stage approach by combining offline ANN-models, with an adaptable/online Kalman-filter to yield a dynamic model.
+\citet{Zaki2013} uses a two-stage approach by combining offline ANN-models, with an adaptable/online Kalman-filter to yield a dynamic model.
 The model is not actually learning in the long term, but is able to adapt to temporal variations in the current travel time on a journey.
@@
 
-\cite{YanjieDuan2016} proposes to use LSTM model for general highway travel time prediction, and do predict multiple steps ahead, but only for a single link at a time, i.e.\ cross link (spacial) correlations are lost.
+Finally \citet{YanjieDuan2016} proposes to use LSTM model for general highway travel time prediction, and do predict multiple steps ahead, but only for a single link at a time, i.e.\ cross link (spacial) correlations are lost.
 
@@
 To allow this mechanism the persistence of information is controlled by three gates: \emph{input gate}, \emph{forget gate}, and \emph{output gate}.
-Each gate yields a state value at time $t$, respectively $i_t$, $f_t$, and $o_t$, along with the cell output, $h_t$, cf.\ \Cref{eq:lstm}.
+Each gate yields a state value at time $t$, respectively $i_t$, $f_t$, and $o_t$, along with the cell output, $h_t$, cf.\ \Cref{eq:lstm}, where $\circ$ denotes the entry-wise product.
 
@@
 \begin{aligned}
-  i_t &= \sigma \left( W_\mathit{xi} x_t + W_\mathit{hi} h_{t-1} + W_{ci} c_{t-1} + b_i \right) \\
-  f_t &= \sigma \left( W_\mathit{xf} x_t + W_\mathit{hf} h_{t-1} + W_\mathit{cf} c_{t-1} + b_f \right) \\
-  c_t &= f_t c_{t-1} + i_t \mathrm{tanh} \left( W_\mathit{xc} x_t + W_\mathit{hc} h_{t-1} + b_c \right) \\
-  o_t &= \sigma \left( W_\mathit{xo} x_t + W_\mathit{ho} h_{t-1} + W_\mathit{co} c_{t} + b_o \right) \\
-  h_t &= o_t \mathrm{tanh} \left( c_t \right)
+  i_t &= \sigma \left( W_\mathit{xi} x_t + W_\mathit{hi} h_{t-1} + W_{ci} \circ c_{t-1} + b_i \right) \\
+  f_t &= \sigma \left( W_\mathit{xf} x_t + W_\mathit{hf} h_{t-1} + W_\mathit{cf} \circ c_{t-1} + b_f \right) \\
+  c_t &= f_t \circ c_{t-1} + i_t \circ \mathrm{tanh} \left( W_\mathit{xc} x_t + W_\mathit{hc} h_{t-1} + b_c \right) \\
+  o_t &= \sigma \left( W_\mathit{xo} x_t + W_\mathit{ho} h_{t-1} + W_\mathit{co} \circ c_{t} + b_o \right) \\
+  h_t &= o_t \circ \mathrm{tanh} \left( c_t \right)
 \end{aligned}
@@
 \end{equation}
+\vspace{.5em}
 
@@
   \centering
-  \includegraphics[scale=.6]{drawings/lstm-peepholes.pdf}
+  \includegraphics[scale=.45]{drawings/lstm-peepholes.pdf}
   \caption{Structure of LSTM cell with peephole.}
@@
 Convolutional Neural Networks~(CNN) on the other hand has been widely used for capturing spacial relationships, e.g.\ importance of neighboring pixels in an image.
-As opposed to fully connected layers, where each unit, $i$, in the layer, has a dedicated weight, $w_{ij}$, for all input, $x_j$, convolutional units are only locally connected and reuses the same weights for producing several outputs.
+As opposed to fully connected layers, where each unit, $i$, in the layer, has a dedicated scalar weight, $w_{ij}$, for all input value, $x_j$, convolutional units are only locally connected and reuses the same weights for producing several outputs.
 Instead of considering the entire input-vector, only a fixed sized window, or \emph{convolution}, around each input is considered.
@@
   \centering
-  \includegraphics[scale=.6]{drawings/conv.pdf}
+  \includegraphics[scale=.45]{drawings/conv.pdf}
   \caption{Application of Convolutional filter onto 1D data.}
@@
 This ensures that the output shape of each convolutional unit will always be identical to the input shape, which is often desirable.
-One of the key benefits of convolutional networks is that the number of weights that needs to be learned is greatly reduced compared to fully connected networks, and that learned patterns transfers across space.
+One of the key benefits of convolutional networks is that the number of weights that needs to be learned is considerably reduced compared to fully connected networks, and that learned patterns can be transfered across space.
 I.e.\ the convolutional filters become feature detectors, that in our case can detect spacial patterns across links, e.g.\ congestion forming, etc.
 
-\cite{ConvLSTM} introduced the novel combination of Convolutional and LSTM layers into a single structure, the \emph{Convolutional LSTM}, or simply \emph{ConvLSTM}.
-Specifically the method applies convolutional filters in the input-to-state and state-to-state transitions the LSTM cf.\ \Cref{eq:convlstm}, where $*$ denotes the convolution operator.
+\citet{ConvLSTM} introduced the novel combination of Convolutional and LSTM layers into a single structure, the \emph{Convolutional LSTM}, or simply \emph{ConvLSTM}.
+Specifically the method applies convolutional filters in the \emph{input-to-state} and \emph{state-to-state} transitions the LSTM cf.\ \Cref{eq:convlstm}, where $*$ denotes the convolution operator.
 
@@
 \begin{aligned}
-  i_t &= \sigma \left( W_\mathit{xi} * x_t + W_\mathit{hi} * h_{t-1} + W_{ci} c_{t-1} + b_i \right) \\
-  f_t &= \sigma \left( W_\mathit{xf} * x_t + W_\mathit{hf} * h_{t-1} + W_\mathit{cf} c_{t-1} + b_f \right) \\
-  c_t &= f_t c_{t-1} + i_t \mathrm{tanh} \left( W_\mathit{xc} * x_t + W_\mathit{hc} * h_{t-1} + b_c \right) \\
-  o_t &= \sigma \left( W_\mathit{xo} * x_t + W_\mathit{ho} * h_{t-1} + W_\mathit{co} c_{t} + b_o \right) \\
-  h_t &= o_t \mathrm{tanh} \left( c_t \right)
+  i_t &= \sigma \left( W_\mathit{xi} * x_t + W_\mathit{hi} * h_{t-1} + W_{ci} \circ c_{t-1} + b_i \right) \\
+  f_t &= \sigma \left( W_\mathit{xf} * x_t + W_\mathit{hf} * h_{t-1} + W_\mathit{cf} \circ c_{t-1} + b_f \right) \\
+  c_t &= f_t \circ c_{t-1} + i_t \circ \mathrm{tanh} \left( W_\mathit{xc} * x_t + W_\mathit{hc} * h_{t-1} + b_c \right) \\
+  o_t &= \sigma \left( W_\mathit{xo} * x_t + W_\mathit{ho} * h_{t-1} + W_\mathit{co} \circ c_{t} + b_o \right) \\
+  h_t &= o_t \circ \mathrm{tanh} \left( c_t \right)
 \end{aligned}
@@
 \end{equation}
+\vspace{.5em}
+
+The output dimensionality of a \emph{ConvLSTM} layer is like the traditional CNN-layer determined by the number of filters applied.
+However \emph{ConvLSTM} requires a total of eight filters for each desired output, i.e.\ four \emph{input-to-state} filters ($W_\mathit{xi}$, $W_\mathit{xf}$, $W_\mathit{xc}$, and $W_\mathit{xo}$) and four \emph{state-to-state} filters ($W_\mathit{hi}$, $W_\mathit{hf}$, $W_\mathit{hc}$, and $W_\mathit{ho}$).
+
+% Hmm, i really would like this to get numbered before fig:data_shape, but cant make it work.
+\afterpage{
+\begin{figure*}[!t]
+  \centering
+  \includegraphics[scale=.75]{drawings/conv_lstm.pdf}
+  \caption{Convolutional LSTM network topology.}
+  \label{fig:ConvLSTM}
+\end{figure*}
+}
+%\newpage
 
@@
 \label{sec:model}
+In this section we present the multi-output, multi-time-step model for bus travel time prediction, which uses the \emph{ConvLSTM} introduced in the previous section.
 
-Input data is arranged in $N$ samples, each with $w$ time steps $t-w+1, \ldots, t$, and each time step with $k$ link travel times $e_1, \ldots, e_k$ cf.~\Cref{fig:data_shape}.
-Likewise is the output arranged with predictions of $j$ time steps ahead, $t+1, \ldots, t+j$.
-Thus the input is a 4D-tensor with shape $(N, w, k, 1)$, and the output a 4D-tensor with shape $(N, j, k, 1)$.
-It is emphasized that each prediction consists of travel time predictions for all links for the next $j$ time steps, i.e.\ multi-output, multi-time-step prediction.
+\subsection{Network topology}
+
+\Cref{fig:ConvLSTM} shows the overall network topology, where blue boxes illustrates \emph{input-to-state} convolutions and yellow \emph{state-to-state} convolutions.
+The network uses an encoder/decoder technique that to a large extent follows \cite{ConvLSTM}, where the encoder-block consists of two \emph{ConvLSTM} layers.
+The result is feed into a decoder, or prediction block, also consisting of two \emph{ConvLSTM} layers.
+The architecture allows unequal $w$ and $k$, e.g.\ predict the next $3$ time-steps based on a window size of $20$.
+
+Thus convolutional filters are applied to each input, at each time-step, to the respective LSTM-cell, and also between LSTM-cells in the state-transition.
+Since the time-steps are one-dimensional~(i.e.\ link travel times across links), the filters are also one-dimensional.
+In each of the two blocks, the \emph{ConvLSTMs} are arranged with filter sizes of respectively $10\times1$ and $5\times1$ for each of the layers in the block.
+This size is used both for the \emph{input-to-state} and \emph{state-to-state} convolutional filters.
+Finally each \emph{ConvLSTM} layer 64~outputs, yielding a total of 512 convolutional filters.
+
+To avoid over-fitting doing training \emph{Dropout} \cite{Dropout} is performed between the \emph{ConvLSTM} layers, and \emph{Batch Normalization} \cite{BatchNorm} are also performed before each \emph{ConvLSTM} layer to ensure reasonable input for activations.
+The dropout factor is adjusted to respectively 20\%, 10\% and 10\%.
+
+Each of the \emph{ConvLSTM} layers uses linear activation functions, and the output from the last layer in the decoder-block is feed into a fully connected (FC) layer using the \emph{ReLU} activation function, which also ensures only positive travel times are predicted.
+
+\subsection{Data preparation}
+We expect link travel times from AVL-systems to be available in a tabular form, where each link travel time measurement has a timestamp, and a reference to the link as illustrated in~\Cref{tab:data}.
+\begin{table}[!ht]
+  \centering
+  \footnotesize
+  \begin{tabular}{llr}
+    Timestamp & Linkref. & Link travel time (s) \\ \hline \hline
+    2017-10-10 00:20:02 & 29848:1254 & 63 \\ \hline
+    2017-10-10 00:21:07 & 1254:1255  & 65 \\ \hline
+    2017-10-10 00:21:51 & 1255:10115 & 44 \\ \hline
+    \vdots & \vdots & \vdots 
+  \end{tabular}
+  \caption{Example of raw travel time measurements.}
+  \label{tab:data}
+\end{table}
+
+For the \emph{ConvLSTM} model to be able to capture the desired spatial-–temporal patterns, the input data must be arranged in a suitable manner: I.e.\ in $N$ samples, each with a window of the $w$ lagging time-steps $t-w+1, \ldots, t$, and each time-step with $u$ link travel times $1, \ldots, u$ cf.~\Cref{fig:data_shape}.
+
+Likewise is the output arranged with $N$ predictions of $k$ time steps ahead, $t+1, \ldots, t+k$.
+Thus the input is a 4D-tensor with shape $(N, w, u, 1)$, and the output a 4D-tensor with shape $(N, k, u, 1)$ -- in both bases the last one refers to the single link travel time for each time-step/link combination.
+It is emphasized that each prediction consists of travel time predictions for all links for the next $k$ time-steps, i.e.\ multi-output, multi-time-step prediction.
+
+The $N$ samples are sampled at a fixed time resolution, since we need a shared time reference across all links.
+\Cref{sec:experiments} elaborates on some of the considerations in choosing a sound resolution.
+
 \begin{figure}[!ht]
   \centering
-  \includegraphics[width=0.8\textwidth]{drawings/data_shape-crop.pdf}
+  \includegraphics[scale=1.1]{drawings/data_shape.pdf}
   \caption{Shapes of the input and output data.}
@@
 
-Obviously the travel time varies throughout the time of the day, and the day of the week due to \emph{recurring congestion}.
-In order to reduce the need for the network to learn this recurring variation, data is normalized to focus on deviations from the normal and expected pattern.
-Travel times are centered with the mean for each link, at teh time of day, and day of week, $\mathit{\bar{x}_{l,\mathit{dow},\mathit{tod}}}$, and scaled with the standard for each link, $\sigma_l$, cf.\ \Cref{eq:normalization}.
+\subsection{Detrending}
+Urban bus travel times varies throughout the time of the day, and the day of the week due to \emph{recurring congestion}.
+In order to reduce the need for the network to learn this recurring variation, link travel for link $\mathit{ln} \in \{ 1,\ldots,u\}$, at time-step $t$, $x_{\mathit{ln},t}$, is normalized to focus on deviations from the normal and expected pattern.
+Travel times are centered with the mean for each link, at the time of day, and day of week, $\mathit{\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}}$, and scaled with the standard deviation for each link, $\sigma_\mathit{ln}$, cf.\ \Cref{eq:normalization}.
 
 \begin{equation}
-  \frac{x - \bar{x}_{l,\mathit{dow},\mathit{tod}}}{\sigma_l}
+  x'_{\mathit{ln},t} = \frac{x_{\mathit{ln},t} - \bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}}{\sigma_\mathit{ln}}
   \label{eq:normalization}
@@
 
-\subsection{Network topology and training}
-\Cref{fig:ConvLSTM} shows the overall network topology, where blue boxes illustrates \emph{input-to-state} convolutions and yellow \emph{state-to-state} convolutions.
-The network uses an encoder/decoder technique that to a large extent follows \cite{ConvLSTM}, where the encoder-block consists of two \emph{ConvLSTM} layers.
-The result is feed into a decoder, or prediction block, also consisting of two \emph{ConvLSTM} layers.
-The architecture allows unequal $w$ and $k$, e.g.\ predict the next $3$ time steps based on a window size of $10$.
+A similar normalization is applied to the predicted travel times, $y_{\mathit{ln},t}$, but only using the historical mean and standard deviation, since the true mean and standard deviation obviously is unavailable in real-time prediction scenarios.
 
-Thus convolutional filters are applied to each input, at each time step to the respective LSTM-cell, and also between LSTM-cells in the state-transition.
-Since the time steps are one-dimensional (link travel times across links), the filters are also one-dimensional.
-In total the 4 layers of Convolutional LSTMs are arranged with filter sizes of $5\times1$ for both \emph{input-to-state} and \emph{state-to-state} filters.
+When calculating the mean and standard deviation it can be beneficial to exclude extreme outliers, since both mean and standard deviation are higly sensitive to such measurements.
+A suggested method is to apply \emph{absolute deviation around the median} (MAD) cf.~\cite{Olewuezi2011} when calculating $\mathit{\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}}$ and $\sigma_\mathit{ln}$.
 
-To avoid over-fitting doing training \emph{Dropout} \cite{Dropout} is performed between the \emph{ConvLSTM} layers, and \emph{Batch Normalization} \cite{BatchNorm} are also performed before each \emph{ConvLSTM} layers to ensure reasonable input for activations.
-The dropout factor is adjusted to respectively 20\%, 10\% and 10\%.
+\subsection{Training}
+The network is trained using the \emph{RMSprop}-algorithm \cite{RMSprop}.
+
+During training the variables $\mathit{\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}}$ and $\sigma_\mathit{ln}$ should be calculated solely based on the training set, to emulated real-world application.
+
+\begin{figure*}[!t]
+    \centering
+    \begin{subfigure}[t]{0.31\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{plots/decycle_5min.pdf}
+        \caption{5 min}
+    \end{subfigure}%
+    ~ 
+    \begin{subfigure}[t]{0.31\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{plots/decycle_15min.pdf}
+        \caption{15 min}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[t]{0.31\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{plots/decycle_30min.pdf}
+        \caption{30 min}
+    \end{subfigure}
+
+    \caption{Examples of travel time for a single link at various time resolutions.}
+    \label{fig:resolutions}
+\end{figure*}
+\newpage
+
+\section{Experiments}
+\label{sec:experiments}
+
+For evaluation the method is applied to a dataset from Copenhagen's public transport authority, ``Movia''.
+The dataset consists of 1,2~M~travel time observations for the ``4A'' bus line in the period January 1 to August 31 2017.
+The data points were collected using the real-time AVL-system installed in all vehicles of the line.
 
@@
   \centering
-  \includegraphics[scale=.8]{drawings/conv_lstm.pdf}
-  \caption{Convolutional LSTM network topology.}
-  \label{fig:ConvLSTM}
-\end{figure}
-
-The network is trained using the \emph{RMSprop}-algorithm \cite{RMSprop}.
-
-\section{Experiments}
-\label{sec:experiments}
-For evaluation the method is applied to a dataset from Copenhagen's public transport authority.
-The dataset consists of 814,710~travel time observations for the ``4A'' bus line in the period January 1 to May 31 2017.
-The data points were  collected using the real-time AVL-system installed in all vehicles of the line.
-The geography of the route is shown in \Cref{fig:4a_map}.
-As the line circles central Copenhagen, it is potentially highly sensitive to congestion to/from the city as it intersects with several large corridors along its route.
-
-TODO: How much used for training, how much for testing?
-
-\begin{figure}[!ht]
-  \centering
-  \includegraphics[width=0.6\textwidth]{drawings/4a.png}
+  \includegraphics[width=0.4\textwidth]{drawings/map.png}
   \caption{Geography of the 4A bus line in Copenhagen.}
@@
 
+The geography of the route is shown in \Cref{fig:4a_map}.
+As the line circles Central Copenhagen, it is potentially highly sensitive to congestion to/from the city as it intersects with several large corridors along its route.
+Southeast of the city center the lines splits in different destination patterns (gray), therefore only the first 32 links are considered in this experiment (red).
+
 \subsection{Data preprocessing}
-Since the travel time measurement contains several extreme outliers due to measurement faults, these were labeled and dropped using the \emph{absolute deviation around the median} (MAD) cf.\ \cite{Olewuezi2011}.
-Data is aggregated into 15-minute time steps and normalized cf.~\Cref{sec:model}.
+Furthermore, to allow predictions for fixed time-steps ahead, data was aggregated at a fixed resolution.
+\Cref{fig:resolutions} shows examples of travel time for a single link and a single day at various time resolutions.
+The red dots are actual measurements, and the blue line the aggregated resolution.
+The choice depends of the \emph{expected} frequency of the line, and is a balance between capturing the details and still having a reasonal number of measurements in each time-step.
+
+For this expirement data was aggregated into 15-minute time-steps and normalized cf.~\Cref{sec:model}.
 
@@
 
+%TODO: How much used for training, how much for testing?
 \subsection{Evaluation}
@@
 
-\begin{figure}
+\begin{figure*}[!ht]
   \centering
-  TODO: Needs to be updated ...
-  \includegraphics[width=0.8\textwidth]{plots/conv_lstm_model_loss.pdf}
-  \caption{Training and validation loss evolution.}
+  \includegraphics[width=0.75\textwidth]{plots/comparison_day.pdf}
+  \caption{}
+  \label{fig:comparison_day}
+\end{figure*}
+
+d
+\begin{figure}[!ht]
+  \centering
+  \includegraphics[width=0.4\textwidth]{plots/conv_lstm_model_loss.pdf}
+  \caption{Training and validation loss evolution. TODO: Needs to be updated ...}
   \label{fig:loss}
@@
 
-\begin{table}[!ht]
+\begin{table*}[!ht]
     \caption{Results of the presented and other popular models}
@@
         Historical average &                & 4.90 & 6.32 & 7.38 \% \\ 
+        \hline 
+        Current Model      & t + 1 (15 min) &  &  &  \% \\ 
+        Current Model      & t + 2 (30 min) &  &  &  \% \\ 
+        Current Model      & t + 3 (45 min) &  &  &  \% \\ 
         \hline 
@@
     \end{tabular}
-\end{table}
+\end{table*}
 
@@
 Further research should be invested in the more rare, but highly impacting deviations, e.g.\ traffic incidents, extreme weather conditions, etc.
+
+\clearpage
 
@@
 
-\newpage
-
-\bibliographystyle{elsarticle-num}
+\section*{References}
+\bibliographystyle{elsarticle-num-names}
 %\bibliography{../references/library}

### 2017-11-02 map/convert.js

@@ -1,0 +1,11 @@
+const puppeteer = require('puppeteer');
+
+(async () => {
+  const browser = await puppeteer.launch();
+  const page = await browser.newPage();
+  await page.setViewport({width: 800, height: 1000});
+  await page.goto('file://C:/Users/ncp/Jupyter/bus-arrival-convlstm/map/map.html');
+  await page.screenshot({path: 'map.png'});
+
+  await browser.close();
+})();

### 2017-11-02 jupyter/common.py
Update common.py

@@ -1,8 +1,6 @@
 import numpy as np
 import pandas as pd
 
-from scipy.stats import rankdata
-
 def prep_data(file, sep = ';', ref_freq = '15min'):
     data = pd.read_csv(file, sep)
 
@@ -16,7 +14,7 @@ def prep_data(file, sep = ';', ref_freq = '15min'):
     
     return data
 
-def fit_scale(data, smooth = 1, ref_freq = '15min'):
+def fit_scale(data, ref_freq = '15min'):
     means = { }
     scales = { }
     low = { }
@@ -29,8 +27,8 @@ def fit_scale(data, smooth = 1, ref_freq = '15min'):
         error = pd.concat([data_link['DowTimeRef'], np.abs(data_link['LinkTravelTime'] - median[data_link['DowTimeRef']].values)], axis = 1)
         mad = 1.4826 * error.groupby('DowTimeRef')['LinkTravelTime'].median()
         
-        _low = median - 5 * mad
-        _upr = median + 5 * mad
+        _low = median - 3 * mad
+        _upr = median + 3 * mad
         mask = (_low[data_link['DowTimeRef']].values < data_link['LinkTravelTime']) & (data_link['LinkTravelTime'] < _upr[data_link['DowTimeRef']].values)
         data_link_no = data_link[mask]
         
@@ -38,17 +36,13 @@ def fit_scale(data, smooth = 1, ref_freq = '15min'):
         means[link_ref] = _mean
         low[link_ref] = _low
         upr[link_ref] = _upr
-        #scales[k] = v_[(low[k] < v_['LinkTravelTime']) & (v_['LinkTravelTime'] < upr[k])]['LinkTravelTime'].std()
-        scales[link_ref] = 1
+        scales[link_ref] = data_link_no['LinkTravelTime'].std()
     
     ix = pd.date_range('1970-01-01', '1970-01-08', freq = ref_freq, closed = 'left')
     means_df = pd.DataFrame(data = means, index = ix).interpolate()
     low_df = pd.DataFrame(data = low, index = ix).interpolate()
     upr_df = pd.DataFrame(data = upr, index = ix).interpolate()
-    
-    if smooth >= 1:
-        means_df = means_df.rolling(window = smooth, center = True).mean()
-        
+            
     # Fill NaNs    
     means_df = means_df.fillna(method='pad').fillna(method='bfill')
     low_df = low_df.fillna(method='pad').fillna(method='bfill')

### 2017-11-02 jupyter/common.py
Update common.py

@@ -48,7 +48,7 @@ def fit_scale(data, ref_freq = '15min'):
     low_df = low_df.fillna(method='pad').fillna(method='bfill')
     upr_df = upr_df.fillna(method='pad').fillna(method='bfill')
     
-    return (means_df, scales, low_df, upr_df)
+    return means_df, scales, low_df, upr_df
 
 def remove_outliers(data, low, upr): 
     _low = low.lookup(data['DowTimeRef'], data['LinkRef'])

### 2017-11-02 paper/paper.tex
Wörk, wörk, wörk.

@@
         \centering
-        \includegraphics[width=\textwidth]{plots/decycle_5min.pdf}
+        \includegraphics[width=\textwidth]{plots/decycle_5min_n.pdf}
         \caption{5 min}
@@
         \centering
-        \includegraphics[width=\textwidth]{plots/decycle_15min.pdf}
+        \includegraphics[width=\textwidth]{plots/decycle_15min_n.pdf}
         \caption{15 min}
@@
         \centering
-        \includegraphics[width=\textwidth]{plots/decycle_30min.pdf}
+        \includegraphics[width=\textwidth]{plots/decycle_30min_n.pdf}
         \caption{30 min}
@@
 
-\subsection{Data preprocessing}
-Furthermore, to allow predictions for fixed time-steps ahead, data was aggregated at a fixed resolution.
+\subsection{Time resolution}
+To allow predictions for fixed time-steps ahead, data is aggregated at a fixed resolution.
 \Cref{fig:resolutions} shows examples of travel time for a single link and a single day at various time resolutions.
-The red dots are actual measurements, and the blue line the aggregated resolution.
+The black dots are actual measurements, and the lines the aggregated link travel time at the given resolution.
 The choice depends of the \emph{expected} frequency of the line, and is a balance between capturing the details and still having a reasonal number of measurements in each time-step.

### 2017-11-07 paper/paper.tex
Wörk, wörk, wörk.

@@
 \documentclass[preprint,11pt,5p,twocolumn]{elsarticle}
+\usepackage[utf8]{inputenc}
 \usepackage{amssymb}
@@
 The goal of this work is to produce precise short-term predictions (e.g.\ $0-1.5$ hours) for link travel time, specifically for bus traffic in urban areas.
-The method is evaluated and compared to other popular approaches for link travel time prediction, and currently available services.
+The method is evaluated and compared to other popular approaches for link travel time prediction, and public available services, e.g.\ Google Traffic.
 
@@
 To allow this mechanism the persistence of information is controlled by three gates: \emph{input gate}, \emph{forget gate}, and \emph{output gate}.
-Each gate yields a state value at time $t$, respectively $i_t$, $f_t$, and $o_t$, along with the cell output, $h_t$, cf.\ \Cref{eq:lstm}, where $\circ$ denotes the entry-wise product.
+Each gate yields a state value at time $t$, respectively $i_t$, $f_t$, and $o_t$, along with the cell output, $h_t$, cf.~\cref{eq:lstm}, where $\circ$ denotes the entry-wise product.
 
@@
 \citet{ConvLSTM} introduced the novel combination of Convolutional and LSTM layers into a single structure, the \emph{Convolutional LSTM}, or simply \emph{ConvLSTM}.
-Specifically the method applies convolutional filters in the \emph{input-to-state} and \emph{state-to-state} transitions the LSTM cf.\ \Cref{eq:convlstm}, where $*$ denotes the convolution operator.
+Specifically the method applies convolutional filters in the \emph{input-to-state} and \emph{state-to-state} transitions the LSTM cf.~\cref{eq:convlstm}, where $*$ denotes the convolution operator.
 
@@
 Likewise is the output arranged with $N$ predictions of $k$ time steps ahead, $t+1, \ldots, t+k$.
-Thus the input is a 4D-tensor with shape $(N, w, u, 1)$, and the output a 4D-tensor with shape $(N, k, u, 1)$ -- in both bases the last one refers to the single link travel time for each time-step/link combination.
+Thus the input is a 4D-tensor, $X$ with shape $(N, w, u, 1)$, and the output, $Y$, a 4D-tensor with shape $(N, k, u, 1)$ -- in both bases the last one refers to the single link travel time for each time-step/link combination.
 It is emphasized that each prediction consists of travel time predictions for all links for the next $k$ time-steps, i.e.\ multi-output, multi-time-step prediction.
@@
 In order to reduce the need for the network to learn this recurring variation, link travel for link $\mathit{ln} \in \{ 1,\ldots,u\}$, at time-step $t$, $x_{\mathit{ln},t}$, is normalized to focus on deviations from the normal and expected pattern.
-Travel times are centered with the mean for each link, at the time of day, and day of week, $\mathit{\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}}$, and scaled with the standard deviation for each link, $\sigma_\mathit{ln}$, cf.\ \Cref{eq:normalization}.
+Travel times are centered with the mean for each link, at the time of day, and day of week, $\mathit{\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}}$, and scaled with the standard deviation for each link, $\sigma_\mathit{ln}$, cf.~\cref{eq:normalization}.
 
@@
 
-\subsection{Training}
-The network is trained using the \emph{RMSprop}-algorithm \cite{RMSprop}.
+\subsection{Implementation and training}
+The proposed network model was implemented in Python using the Keras Framework \cite{Keras}, and trained using the \emph{RMSprop}-algorithm \cite{RMSprop}.
 
@@
 
-    \caption{Examples of travel time for a single link at various time resolutions.}
+    \caption{Examples of travel time for a single link over a single day, at various time resolutions.}
     \label{fig:resolutions}
@@
 For evaluation the method is applied to a dataset from Copenhagen's public transport authority, ``Movia''.
-The dataset consists of 1,2~M~travel time observations for the ``4A'' bus line in the period January 1 to August 31 2017.
-The data points were collected using the real-time AVL-system installed in all vehicles of the line.
+The dataset consists of 1,2~M~travel time observations for the ``4A'' bus line in the period May to October 2017.
+The data points were collected using the real-time AVL-system installed in every vehicle servicing the line.
 
@@
 As the line circles Central Copenhagen, it is potentially highly sensitive to congestion to/from the city as it intersects with several large corridors along its route.
-Southeast of the city center the lines splits in different destination patterns (gray), therefore only the first 32 links are considered in this experiment (red).
+Southeast of the city center the line splits in different destination patterns (gray), therefore only the first 32 links are considered in this experiment (red).
 
 \subsection{Time resolution}
-To allow predictions for fixed time-steps ahead, data is aggregated at a fixed resolution.
-\Cref{fig:resolutions} shows examples of travel time for a single link and a single day at various time resolutions.
-The black dots are actual measurements, and the lines the aggregated link travel time at the given resolution.
-The choice depends of the \emph{expected} frequency of the line, and is a balance between capturing the details and still having a reasonal number of measurements in each time-step.
+To allow predictions for fixed time-steps ahead, data is aggregated at a fixed time resolution.
+\Cref{fig:resolutions} shows examples of travel time for a single link over a single day at various time resolutions.
+The black dots are actual measurements, and the lines the aggregated mean link travel time at the given resolution.
+The choice depends of the \emph{expected} frequency of the line, and is a balance between capturing the details and still having a reasonal number of measurements in each time-step to avoid over-fitting.
 
-For this expirement data was aggregated into 15-minute time-steps and normalized cf.~\Cref{sec:model}.
-
-The implementation of the proposed model was done in Python using Keras \cite{Keras}.
+For this experiment data was aggregated into 15-minute time-steps and normalized cf.~\Cref{sec:model}.
+This resolution were chosen based on a measured mean \emph{headway} of $7.5$ minutes between 06:00 and 22:00, i.e.\ the time between two vehicles during daytime.
 
@@
 \subsection{Evaluation}
-In order to evaluate the presented model and comparisons, the following measures is used: \emph{mean absolute error} (MAE), \emph{root mean square error} (RMSE), and \emph{mean absolute percentage error} (MAPE) cf.~\Cref{eq:mae,eq:rmse,eq:mape}, where $X_i$ is the true travel time for sample $i$ and $\widehat{X}_i$ is the predicted travel time.
+In order to evaluate the presented model and comparisons, the following measures is used: \emph{mean absolute error} (MAE), \emph{root mean square error} (RMSE), and \emph{mean absolute percentage error} (MAPE) cf.~\cref{eq:mae,eq:rmse,eq:mape}, where $Y_i$ is the true link travel times for sample $i$ and $\widehat{Y}_i$ is the predicted travel time.
 
 \begin{equation}
-    \textrm{MAE}(X, \widehat{X}) = \frac{\sum_{i = 1}^{n} \left| X_i - \widehat{X}_i \right| }{n}
+    \textrm{MAE}(Y, \widehat{Y}) = \frac{\sum_{i = 1}^{N} \left| Y_i - \widehat{Y}_i \right| }{N}
     \label{eq:mae}
@@
 \begin{equation}
-    \textrm{RMSE}(X, \widehat{X}) = \sqrt{\frac{\sum_{i = 1}^{n} \left(X_i - \widehat{X}_i \right)^2}{n}}
+    \textrm{RMSE}(Y, \widehat{Y}) = \sqrt{\frac{\sum_{i = 1}^{N} \left(Y_i - \widehat{Y}_i \right)^2}{N}}
     \label{eq:rmse}
@@
 \begin{equation}
-    \textrm{MAPE}(X, \widehat{X}) = \frac{1}{n} \sum_{i = 1}^{n} \left| \frac{X_i - \widehat{X}_i}{X_i} \right| 
+    \textrm{MAPE}(Y, \widehat{Y}) = \frac{1}{N} \sum_{i = 1}^{N} \left| \frac{Y_i - \widehat{Y}_i}{Y_i} \right| 
     \label{eq:mape} 
 \end{equation}
+\vspace{.5em}
+
+\begin{table*}[!t]
+    \center
+    \begin{tabular}{ll|rrr}
+        Model & Time ahead & RMSE (min) & MAE (min) & MAPE (\%) \\
+        \hline         
+        \hline
+        Historical average &                & 4.35 & 3.23 & 6.51 \% \\ 
+        \hline 
+        Current model      & t + 1 (15 min) & 4.92 & 3.90 & 8.05 \% \\
+                           & t + 2 (30 min) & 4.91 & 3.46 & 6.82 \% \\
+                           & t + 3 (45 min) & 5.47 & 4.15 & 8.68 \% \\
+        \hline 
+        Pure LSTM          & t + 1 (15 min) & 3.48 & 2.48 & 5.02 \% \\
+                           & t + 2 (30 min) & 3.56 & 2.51 & 5.08 \% \\
+                           & t + 3 (45 min) & 3.68 & 2.62 & 5.34 \% \\
+        \hline 
+        Google Traffic     & t + 1 (15 min) & 3.67 & 2.96 & 6.32 \% \\ 
+        \hline 
+        ConvLSTM           & t + 1 (15 min) & 2.66 & 1.99 & 4.19 \% \\
+                           & t + 2 (30 min) & 2.89 & 2.11 & 4.44 \% \\
+                           & t + 3 (45 min) & 3.11 & 2.27 & 4.75 \% \\
+        \hline 
+    \end{tabular}
+    \caption{Results of the presented and other popular models}
+    \label{tab:results}
+\end{table*}
+
+Since the multi-output, multi-time-step model predicts link travel times for all $u$ links for the next $k$ time steps, $Y_i$ and $\widehat{Y}_i$ are both $u \times k $ matrices.
+To allow a clear comparison, we reduce $Y_i$ and $\widehat{Y}_i$ by summing over all links cf.~\cref{eq:total}.
+This is equivalent of predicting the total travel time of all 32 links, and follows the initial approach for arrival/departure time prediction by accumulating link travel times.
+\begin{align}
+    Y_i' = \sum_{\mathit{ln} = 1}^{u} Y_{i,\mathit{ln}} & & \widehat{Y}_i' = \sum_{\mathit{ln} = 1}^{u} \widehat{Y}_{i,\mathit{ln}}
+    \label{eq:total} 
+\end{align}
+
+The model is trained on the prepared data using a sliding window approach to simulate real-world conditions where real-time travel time measurements arrives as a continuously data stream.
+We use 23 weeks of data for training, and one week of data for testing.
 
@@
 \label{sec:results}
-The model is trained on the prepared data using a sliding window approach to simulate real-world conditions where real-time travel time measurements arrives as a continuously data stream.
-\Cref{fig:loss} shows the evolution of training and validation loss for each epoch for such a window, where the loss is the MAE if the individual links.
+The performance of our \emph{ConvLSTM} model for link travel time prediction is compared several other models and services using the presented evaluation:
+\begin{enumerate}
+   \item a naïve historical average model, i.e.\ equivalent of just predicting the normalized value, $\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}$ 
+   \item the current traffic prediction model deployed by ``Movia''.
+   \item a pure LSTM-model for link travel time prediction, i.e.\ without applying convolutional filters in state transitions.
+   \item travel time predictions from Google Traffic (part of Google Maps).
+\end{enumerate}
 
-\begin{figure*}[!ht]
+\begin{figure*}[!t]
   \centering
-  \includegraphics[width=0.75\textwidth]{plots/comparison_day.pdf}
+  \includegraphics[width=\textwidth]{plots/comparison_day.pdf}
   \caption{}
@@
 
-d
-\begin{figure}[!ht]
-  \centering
-  \includegraphics[width=0.4\textwidth]{plots/conv_lstm_model_loss.pdf}
-  \caption{Training and validation loss evolution. TODO: Needs to be updated ...}
-  \label{fig:loss}
-\end{figure}
+\Cref{tab:results} shows the overall performance of the presented and the compared models.
+Predictions are limited to daytime, i.e.\ between 06:00 and 22:00 and accumulated downstream on journey level to simulate for use in real-time bus arrival/departure time prediction cf~\cref{eq:total}.
 
-\Cref{tab:results} shows the results of the presented method and compares to two other methods: 1) a naïve historical average model, i.e.
-equivalent of just prediction the normalized value, $\bar{x}_{l,\mathit{dow},\mathit{tod}}$ and 2) a pure LSTM model as proposed by \cite{YanjieDuan2016}.
-Predictions are accumulated downstream on journey level to simulate for use in real-time bus arrival/departure time prediction.
+Before going into a direct comparison, it is important to understand some aspects of the compared models, and how measurements were collected.
 
+\subsection{Historical average}
+The performance of the historical average is independent with respect to the time steps ahead it predicts, as it just represents a weekly cycle of mean link travel times.
+
+\subsection{Current model}
+Measurements for the currently deployed bus prediction model was collected at a 5-minute frequency using a non-publicly accessible endpoint at the transport authority.
+The model is based on an historical average model, but has rule based mechanism on top, which can override or adjust the historical link travel times.
+For instance it will assume that a delayed vehicle can recover (partially or completely) from its delay by traversing links faster.
+Of course such an assumption can be problematic in an urban area with lots of external traffic effects.
+
+\subsection{Pure LSTM}
+The pure LSTM-model for link travel time prediction is similar to the model proposed by \citet{YanjieDuan2016}.
+The model was trained on the exact same dataset as the \emph{ConvLSTM} model, and has a similar architecture, but without the convolutional filters.
+\newpage
+
+\subsection{Google Traffic}
+Measurements from the Google Traffic model were collected using the Google Maps Distance Matrix API~\citep{GMaps_DistanceMatrixAPI}.
+Google uses crowd-sourced road congestion data collected from smart-phones with the \emph{Google Maps App} installed \citep{GMaps_Crowdsourcing}.
+While the exact model powering the service is not publicly described in details, the documentation states that ``estimates of travel time is given (with respect to) what is known about both historical traffic conditions and live traffic''.
+And further, ``that live traffic becomes more important the closer the departure time is to now''.
+
+Because there is a limit on the number of request you can make freely to the API over a 24-hour period, it has only been possible to collect link travel times for the $t + 1$ time step (i.e.\ next 15 minutes).
+
+Another important aspect is that the Google Traffic model is primary designed for estimating car driving times, and can thus be biased away from bus link travel times used in this experiment.
+
+\subsection{Comparison}
 From the table it is found that the Convolutional LSTM performs better than the compared methods.
 Although the advantage might seem small it is emphasized that evaluation measurements are averaging their response, and thus the increased accuracy can be much higher on individual journeys, especially if they experienced very irregular travel times.
-
-\begin{table*}[!ht]
-    \caption{Results of the presented and other popular models}
-    \label{tab:results}
-    \center
-    \begin{tabular}{ll|rrr}
-        Model & Time ahead & MAE (min) & RMSE (min) & MAPE (\%) \\
-        \hline 
-        Historical average &                & 4.90 & 6.32 & 7.38 \% \\ 
-        \hline 
-        Current Model      & t + 1 (15 min) &  &  &  \% \\ 
-        Current Model      & t + 2 (30 min) &  &  &  \% \\ 
-        Current Model      & t + 3 (45 min) &  &  &  \% \\ 
-        \hline 
-        Pure LSTM          & t + 1 (15 min) & 3.71 & 3.71 & 5.87 \% \\
-        Pure LSTM          & t + 2 (30 min) & 4.31 & 5.63 & 6.63 \% \\
-        Pure LSTM          & t + 3 (45 min) & 4.97 & 6.41 & 7.52 \% \\
-        \hline 
-        Convolutional LSTM & t + 1 (15 min) & 2.96 & 3.94 & 5.10 \% \\
-        Convolutional LSTM & t + 2 (30 min) & 3.04 & 4.08 & 5.21 \% \\
-        Convolutional LSTM & t + 3 (45 min) & 3.17 & 4.24 & 5.37 \% \\
-        \hline 
-    \end{tabular}
-\end{table*}
 

### 2017-11-21 paper/paper.tex
Finish paper for review.

@@
 \usepackage{amssymb}
+\usepackage{bm}
 \usepackage{mathtools}
+\usepackage{xfrac}
 \usepackage{graphicx}
@@
 \usepackage{afterpage}
+
+\newcommand{\matr}[1]{\mathbf{#1}}
 
@@
 }
+
+\let\oldhref\href
+\renewcommand{\href}[2]{\oldhref{#1}{\hbox{#2}}}
 \makeatother
@@
 \title{Multi-output Bus Travel Time Prediction\\
-Using Convolutional LSTM Neural Network}
+with Convolutional LSTM Neural Networks}
 
@@
 
-\author{Niklas Christoffer Petersen}
+\author{Niklas Christoffer Petersen\\[.25em]\footnotesize{niklch@dtu.dk}}
 
 \address{
+  \vspace{.5em}
+  \begin{minipage}{.5\textwidth}
+    \center
     Department of Management Engineering \\
-  Technical University of Denmark, 2800 Kongens Lyngby \\
-  niklch@dtu.dk
+    Technical University of Denmark, 2800 Kongens Lyngby 
+  \end{minipage}% This must go next to `\end{minipage}`
+  \begin{minipage}{.5\textwidth}
+    \center
+    Public Transport Movia \\
+    Gammel Køge Landevej 3, 2500 Valby 
+  \end{minipage}
+  \vspace{-2.5em}
 }
@@
 \begin{abstract}
-Accurate and reliable predictions for travel times in a public transport networks is essential for delivering an attractive service.
+Accurate and reliable predictions for travel times in public transport networks is essential for delivering an attractive service.
 This paper presents a multi-output, multi-time-step, deep neural network for bus travel time prediction using Convolutional and Long short-term memory (LSTM) layers.
-The method is evaluated and compared to other popular approaches for link travel time prediction, and currently available services.
+The method is empirically evaluated and compared to other popular approaches for link travel time prediction, and currently available services.
+We find that the proposed model significantly outperforms all the other methods, and is able to detect small irregular peeks in bus travel time very quickly.
 \end{abstract}
@@
 %% keywords here, in the form: keyword \sep keyword
-Bus Travel Time Prediction \sep Intelligent Transport Systems \sep Convolutional Neural Network (CNN) \sep Long short-term memory (LSTM)
+Bus Travel Time Prediction \sep Intelligent Transport Systems \sep Convolutional Neural Network (CNN) \sep Long short-term memory (LSTM), Deep Learning.
 \end{keyword}
@@
 \section{Introduction}
-Public transport authorities has long found that GPS trajectory data from already deployed \emph{Automatic Vehicle Location}-systems (AVL) can be used in \emph{Intelligent Transport Systems} (ITS) \cite{Tcrp48}.
-Examples include real-time traffic information for passengers, e.g.\ departure boards, where studies has shown that reliable real-time information at bus stops has a statistical significant dampening effect on the perceived waiting time \cite{Fan2016}.
+Public transport authorities have long found that GPS trajectory data from already deployed \emph{Automatic Vehicle Location}-systems (AVL) can be used in \emph{Intelligent Transport Systems} (ITS) \cite{Tcrp48}.
+Examples include real-time traffic information for passengers, e.g.\ smartphone travel planner apps, in-vehicle infotainment screens and departure boards.
+Studies has shown that reliable real-time information at bus stops has a statistical significant dampening effect on the perceived waiting time \cite{Fan2016}.
 
-Besides real-time passenger information, robust arrival- and departure time predictions are necessary for operating more sophisticated ITS applications successfully, e.g.\ demand-adaptive transit systems, and \emph{connection insurance} between different public transport services.
-Arrival/departure time prediction is commonly approached as a specialization of travel time prediction, where the predicted travel time is simply accumulated downstream the route to yield the arrival/departure time predictions at each stop point of the rest of the current journey.
+Besides real-time passenger information, robust arrival- and departure time predictions are necessary for operating more sophisticated ITS applications successfully, e.g.
+\emph{connection assurance} between low frequent public transport services, where accurate arrival times are needed to decide if one of the services should wait on the other.
+Another example is \emph{demand-adaptive transit systems}, where prolonged travel time should be taken into account because of the influence on both vehicle and driver utilization, and thus the efficiency of the transit system.
+Arrival/departure time prediction is commonly approached as a specialization of travel time prediction as illustrated in \Cref{fig:bus-arrival-links}.
+The predicted travel time for each link is simply accumulated downstream the route to yield the arrival/departure time predictions at each stop point of the rest of the current journey.
 Besides the link travel time, estimations of dwell time (i.e.\ when a bus is holding at a stop point) are also accumulated downstream.
 
+\begin{figure}[!ht]
+  \center  
+  \includegraphics[width=\columnwidth]{drawings/bus-arrival-links-narrow.pdf}
+  \caption{Arrival- and departure using link travel time.}
+  \label{fig:bus-arrival-links}
+\end{figure}
+
 Producing precise bus travel time predictions in areas with little external influence, e.g.\ rural areas, can be solved to a large extent with historical averaging or simple regression methods \cite{Williams2003,Altinkaya2013}.
-The problem becomes much more complex in urban areas where congestion, events, road-works, weather, etc.\ highly influences the traffic flow and passenger demand.
+The problem becomes much more complex in urban areas where congestion, events, road-works, weather, etc.\ highly influence the traffic flow and passenger demand.
+As on-board GPS and AVL-systems has become more affordable and common, data has both grown in coverage, i.e.\ number of vehicles with AVL installed, and frequency, i.e.\ number of GPS-positions collected per vehicle.
 
-In this paper we presents a multi-output, multi-time-step deep neural network for bus travel time prediction using a combination of \emph{Convolutional} and \emph{Long short-term memory} (LSTM) \cite{Lstm1,Lstm2} layers.
+In this paper we present a multi-output, multi-time-step deep neural network for bus travel time prediction using a combination of \emph{Convolutional} and \emph{Long short-term memory} (LSTM) \cite{Lstm1,Lstm2} layers.
 The goal of this work is to produce precise short-term predictions (e.g.\ $0-1.5$ hours) for link travel time, specifically for bus traffic in urban areas.
-The method is evaluated and compared to other popular approaches for link travel time prediction, and public available services, e.g.\ Google Traffic.
+The method is empirically evaluated and compared to other popular approaches for link travel time prediction, including the currently model deployed in production by the public transport authority for the Greater Copenhagen Area: \emph{Movia}.
+Furthermore the method is compared to Google Traffic, a popular online service for travel time prediction.
 
-The paper is structured in the following manner: In the next section related work and literature is reviewed.
+This paper is structured in the following manner: In the next section, related work and literature are reviewed.
 \Cref{sec:convlstm} introduces Convolutional LSTM neural networks in general and in \Cref{sec:model} we present the proposed multi-output model in more details, including e.g.\ network topology, etc.
@@
 
-\subsection{Literature review}
-Early approaches presents historical averaging models \cite{Dailey1999,Sun2007}, and linear regression \cite{Patnaik2004}.
-Recent research presents this type of model only for comparison purpose, and the models are in all cases outperformed by the compared models \cite{Shalaby2004,Jeong2005}.
-Kalman-filters, which by its capabilities of maintaining state between predictions, has shown interest from several studies, either as an independent model \cite{Chen2001,Shalaby2004} or in combination with other models \cite{Zaki2013,Bai2015}.
+\section{Literature review}
+Bus link travel time prediction has been explored in research as GPS and AVL data has become increasingly available.
+The problem overlaps with other research areas such as general traffic flow and speed estimation, but also has unique constraints and opportunities that follows from servicing a fixed route.
+The improvement in computational power in the recent decades has gradually allowed more complex link travel time models with increased precision.
 
-Both the different regression style models and the Kalman-filter models still has limited options for capturing fluctuations in travel and dwell time in a metropolitan bus system, since they to a large extent still is averaging and smoothing their response.
+Early approaches for bus travel time prediction rely on historical averaging models \cite{Dailey1999,Sun2007}, and linear regression \cite{Patnaik2004}.
+Recent research presents this type of model only for comparison purposes, and the models are in all cases outperformed by the proposed models \cite{Shalaby2004,Jeong2005}.
+A historical averaging model will only slowly converge to changes in the travel time, which of cause is undesired with short, but highly impacting, external influences (e.g.\ a traffic incident or large event).
+
+Kalman-filters, which by its capabilities of maintaining state between predictions, has shown interest from several studies for capturing \emph{changes} in the travel time, either as an independent model \cite{Chen2001,Shalaby2004} or in combination with other models \cite{Yo2010,Bai2015}.
+The Kalman-filter models still have limited options for capturing and forecasting the complex dynamics of travel and dwell time in a metropolitan bus system.
 E.g., the Kalman-filter's state is only directly accessible for the leading time-step, and thus not capable of finding long-distance patterns spanning over several links and/or over several time-steps.
-This is substantiated by \cite{Lin2013} and \cite{Kumar2014} that finds artificial neural networks (ANN) outperforms Kalman-filter models.
+This is substantiated by \cite{Lin2013} and \cite{Kumar2014}, who find artificial neural networks (ANN) to outperform Kalman-filter models.
+However, the computational challenges of fully connected ANNs is also limiting the depth of the network and thus how complex patterns it can learn to recognize.
+This has sparked the interest for studying composite or hybrid models.
+\citet{Bai2015} uses a two-stage approach by combining offline ANN-models, with an adaptable/online Kalman-filter to yield a dynamic model.
+The model is not actually learning in the long term, but is able to adapt to temporal variations in the current travel time on a journey, and is still not able to recognize long distance patterns.
+The proposed model in this work uses \emph{Long short-term memory} cells (LSTM), and differs from existing research in bus link travel time prediction by combining the capability for persisting state-space over multiple time-steps, while allowing the deep neural network to be feasibly trainable.
 
-The computational challenges of pure ANN approaches i.a.\ has sparked the interest for studying composite or hybrid models.
-\citet{Zaki2013} uses a two-stage approach by combining offline ANN-models, with an adaptable/online Kalman-filter to yield a dynamic model.
-The model is not actually learning in the long term, but is able to adapt to temporal variations in the current travel time on a journey.
-
-Some recent research recognizes that several routes can benefit from each other's predictions if they share some partial route segment, e.g.\ \cite{Yu2011,Gal2014,Bai2015}, however none of these approaches consider cross temporal correlations between different route segments, and only uses a small window for correlation with upstream links (e.g.\ max.
-3 links).
-
-Finally \citet{YanjieDuan2016} proposes to use LSTM model for general highway travel time prediction, and do predict multiple steps ahead, but only for a single link at a time, i.e.\ cross link (spacial) correlations are lost.
+Some recent research recognizes that several routes can benefit from each other's predictions if they share some partial route segment, e.g.\ \cite{Yu2011,Gal2014,Bai2015}, however none of these approaches consider cross temporal correlations between different route segments, and only use a small window for correlation with upstream links (e.g.\ max.\ 3 links).
+Likewise, \citet{YanjieDuan2016} proposes to use a LSTM model for general highway travel time prediction, and do predict multiple time-steps ahead, but only for a single link at a time, i.e.\ cross link (spatial) correlations are lost.
+In contrast the combination of both \emph{LSTM} cells and \emph{Convolutional} filters for bus travel time prediction, proposed in this paper, allows the learned patterns to generalize beyond a single link and time, i.e.\ multi-output and multi-time-step.
+Furthermore this reduces the computational complexity by magnitudes compared to fully connected ANNs capable of capturing similar complex patterns.
 
@@
 A Long short-term memory (LSTM) neural network is a special type of Recurrent Neural Network~(RNN) which has proven robust for capturing long-distance dependencies \cite{Lstm1,Lstm2}.
-The important feature of a LSTM network is its capability to persist cell state, $c_t$, from previous observations across sequences of input (e.g.\ time), but also eliminate information which is considered unimportant.
+The important feature of a LSTM network is its capability to persist cell state, $\matr{c}_t$, from previous observations across sequences of input (e.g.\ time), but also eliminate information which is considered irrelevant.
 To allow this mechanism the persistence of information is controlled by three gates: \emph{input gate}, \emph{forget gate}, and \emph{output gate}.
-Each gate yields a state value at time $t$, respectively $i_t$, $f_t$, and $o_t$, along with the cell output, $h_t$, cf.~\cref{eq:lstm}, where $\circ$ denotes the entry-wise product.
+Each gate yields a state value at time $t$, respectively $\matr{i}_t$, $\matr{f}_t$, and $\matr{o}_t$, along with the cell output, $\matr{h}_t$, cf.~\cref{eq:lstm}, where $\circ$ denotes the entry-wise product.
 
@@
 \begin{aligned}
-  i_t &= \sigma \left( W_\mathit{xi} x_t + W_\mathit{hi} h_{t-1} + W_{ci} \circ c_{t-1} + b_i \right) \\
-  f_t &= \sigma \left( W_\mathit{xf} x_t + W_\mathit{hf} h_{t-1} + W_\mathit{cf} \circ c_{t-1} + b_f \right) \\
-  c_t &= f_t \circ c_{t-1} + i_t \circ \mathrm{tanh} \left( W_\mathit{xc} x_t + W_\mathit{hc} h_{t-1} + b_c \right) \\
-  o_t &= \sigma \left( W_\mathit{xo} x_t + W_\mathit{ho} h_{t-1} + W_\mathit{co} \circ c_{t} + b_o \right) \\
-  h_t &= o_t \circ \mathrm{tanh} \left( c_t \right)
+  \matr{i}_t &= \sigma \left( \matr{W}^i \matr{x}_t + \matr{R}^i \matr{h}_{t-1} + \matr{U}^i \circ \matr{c}_{t-1} + \matr{b}^i \right) \\
+  \matr{f}_t &= \sigma \left( \matr{W}^f \matr{x}_t + \matr{R}^f \matr{h}_{t-1} + \matr{U}^f \circ \matr{c}_{t-1} + \matr{b}^f \right) \\
+  \matr{c}_t &= \matr{f}_t \circ \matr{c}_{t-1} + \matr{i}_t \circ \mathrm{tanh} \left( \matr{W}^c \matr{x}_t + \matr{R}^c \matr{h}_{t-1} + \matr{b}_c \right) \\
+  \matr{o}_t &= \sigma \left( \matr{W}^\mathit{o} \matr{x}_t + \matr{R}^\mathit{o} \matr{h}_{t-1} + \matr{U}^\mathit{o} \circ \matr{c}_{t} + \matr{b}^o \right) \\
+  \matr{h}_t &= \matr{o}_t \circ \mathrm{tanh} \left( \matr{c}_t \right)
 \end{aligned}
@@
 Is has especially grown popular for predicting time series using methods evolved from \cite{LstmTs}, where fixed-length windows of time-series are generated and feed into a LSTM network.
-Multiple LSTMs can be stacked such more complex patterns of sequential information (e.g.\ temporal patterns) can be learned.
+Multiple LSTMs can be stacked such that more complex patterns of sequential information (e.g.\ temporal patterns) can be learned.
 
@@
 
-Convolutional Neural Networks~(CNN) on the other hand has been widely used for capturing spacial relationships, e.g.\ importance of neighboring pixels in an image.
-As opposed to fully connected layers, where each unit, $i$, in the layer, has a dedicated scalar weight, $w_{ij}$, for all input value, $x_j$, convolutional units are only locally connected and reuses the same weights for producing several outputs.
-Instead of considering the entire input-vector, only a fixed sized window, or \emph{convolution}, around each input is considered.
+Convolutional Neural Networks~(CNN) on the other hand have been widely used for capturing spacial relationships, e.g.\ importance of neighboring pixels in an image.
+As opposed to fully connected layers, where each unit, $i$, in the layer, has a dedicated scalar weight, $w_{ij}$, for all input value, $x_j$, convolutional units are only locally connected and reuse the same weights for producing several outputs.
+Instead of considering the entire input-vector, only a fixed-size window, or \emph{convolution}, around each input is considered.
 The weights are therefore referred to as the \emph{filters} or \emph{kernels} of the layer.
@@
 One of the key benefits of convolutional networks is that the number of weights that needs to be learned is considerably reduced compared to fully connected networks, and that learned patterns can be transfered across space.
-I.e.\ the convolutional filters become feature detectors, that in our case can detect spacial patterns across links, e.g.\ congestion forming, etc.
+I.e.\ the convolutional filters become feature detectors, that in our case can detect spatial patterns across links, e.g.\ congestion forming, etc.
 
 \citet{ConvLSTM} introduced the novel combination of Convolutional and LSTM layers into a single structure, the \emph{Convolutional LSTM}, or simply \emph{ConvLSTM}.
-Specifically the method applies convolutional filters in the \emph{input-to-state} and \emph{state-to-state} transitions the LSTM cf.~\cref{eq:convlstm}, where $*$ denotes the convolution operator.
-
+Specifically the method applies convolutional filters in the \emph{input-to-state} and \emph{state-to-state} transitions of the LSTM cf.~\cref{eq:convlstm}, where $*$ denotes the convolution operator.
 \begin{equation}
 \begin{aligned}
-  i_t &= \sigma \left( W_\mathit{xi} * x_t + W_\mathit{hi} * h_{t-1} + W_{ci} \circ c_{t-1} + b_i \right) \\
-  f_t &= \sigma \left( W_\mathit{xf} * x_t + W_\mathit{hf} * h_{t-1} + W_\mathit{cf} \circ c_{t-1} + b_f \right) \\
-  c_t &= f_t \circ c_{t-1} + i_t \circ \mathrm{tanh} \left( W_\mathit{xc} * x_t + W_\mathit{hc} * h_{t-1} + b_c \right) \\
-  o_t &= \sigma \left( W_\mathit{xo} * x_t + W_\mathit{ho} * h_{t-1} + W_\mathit{co} \circ c_{t} + b_o \right) \\
-  h_t &= o_t \circ \mathrm{tanh} \left( c_t \right)
+  \matr{i}_t &= \sigma \left( \matr{W}^i * \matr{x}_t + \matr{R}^i * \matr{h}_{t-1} + \matr{U}^i \circ \matr{c}_{t-1} + \matr{b}^i \right) \\
+  \matr{f}_t &= \sigma \left( \matr{W}^f * \matr{x}_t + \matr{R}^f * \matr{h}_{t-1} + \matr{U}^f \circ \matr{c}_{t-1} + \matr{b}^f \right) \\
+  \matr{c}_t &= \matr{f}_t \circ \matr{c}_{t-1} + \matr{i}_t \circ \mathrm{tanh} \left( \matr{W}^c * \matr{x}_t + \matr{R}^c * \matr{h}_{t-1} + \matr{b}_c \right) \\
+  \matr{o}_t &= \sigma \left( \matr{W}^\mathit{o} * \matr{x}_t + \matr{R}^\mathit{o} * \matr{h}_{t-1} + \matr{U}^\mathit{o} \circ \matr{c}_{t} + \matr{b}^o \right) \\
+  \matr{h}_t &= \matr{o}_t \circ \mathrm{tanh} \left( \matr{c}_t \right)
 \end{aligned}
@@
 
-The output dimensionality of a \emph{ConvLSTM} layer is like the traditional CNN-layer determined by the number of filters applied.
-However \emph{ConvLSTM} requires a total of eight filters for each desired output, i.e.\ four \emph{input-to-state} filters ($W_\mathit{xi}$, $W_\mathit{xf}$, $W_\mathit{xc}$, and $W_\mathit{xo}$) and four \emph{state-to-state} filters ($W_\mathit{hi}$, $W_\mathit{hf}$, $W_\mathit{hc}$, and $W_\mathit{ho}$).
-
-% Hmm, i really would like this to get numbered before fig:data_shape, but cant make it work.
-\afterpage{
 \begin{figure*}[!t]
@@
 \end{figure*}
-}
-%\newpage
+
+The output dimensionality of a \emph{ConvLSTM} layer is like the traditional CNN-layer determined by the number of filters applied.
+However, \emph{ConvLSTMs} require a total of eight filters for each desired output, i.e.\ four \emph{input-to-state} filters ($\matr{W}^i$, $\matr{W}^f$, $\matr{W}^c$, and $\matr{W}^o$) and four \emph{state-to-state} filters ($\matr{R}^i$, $\matr{R}^f$, $\matr{R}^c$, and $\matr{R}^o$).
+Still it is important to emphasize that application of convolutional filters to the LSTM model greatly reduces the number of weights that needs to be learned, compared to a \emph{pure LSTM} approach.
+This allows for even deeper networks.
+\newpage
 
@@
 The network uses an encoder/decoder technique that to a large extent follows \cite{ConvLSTM}, where the encoder-block consists of two \emph{ConvLSTM} layers.
-The result is feed into a decoder, or prediction block, also consisting of two \emph{ConvLSTM} layers.
+The result is fed into a decoder, or prediction block, also consisting of two \emph{ConvLSTM} layers.
 The architecture allows unequal $w$ and $k$, e.g.\ predict the next $3$ time-steps based on a window size of $20$.
@@
 
-To avoid over-fitting doing training \emph{Dropout} \cite{Dropout} is performed between the \emph{ConvLSTM} layers, and \emph{Batch Normalization} \cite{BatchNorm} are also performed before each \emph{ConvLSTM} layer to ensure reasonable input for activations.
-The dropout factor is adjusted to respectively 20\%, 10\% and 10\%.
+To avoid over-fitting during training \emph{Dropout} \cite{Dropout} is performed between the \emph{ConvLSTM} layers, and \emph{Batch Normalization} \cite{BatchNorm} is also performed before each \emph{ConvLSTM} layer to ensure reasonable inputs for the activations.
+The dropout probability is adjusted to respectively 20\%, 10\% and 10\%.
 
-Each of the \emph{ConvLSTM} layers uses linear activation functions, and the output from the last layer in the decoder-block is feed into a fully connected (FC) layer using the \emph{ReLU} activation function, which also ensures only positive travel times are predicted.
+Each of the \emph{ConvLSTM} layers uses linear activation functions, and the output from the last layer in the decoder-block is fed into a fully connected (FC) layer using the \emph{ReLU} activation function, which also ensures that only positive travel times are predicted.
 
@@
 
-Likewise is the output arranged with $N$ predictions of $k$ time steps ahead, $t+1, \ldots, t+k$.
-Thus the input is a 4D-tensor, $X$ with shape $(N, w, u, 1)$, and the output, $Y$, a 4D-tensor with shape $(N, k, u, 1)$ -- in both bases the last one refers to the single link travel time for each time-step/link combination.
+Likewise, the output is arranged with $N$ predictions of $k$ time steps ahead, $t+1, \ldots, t+k$.
+Thus the input is a 4D-tensor, $\matr{X}$ with dimensionality $N \times w \times u \times 1$, and the output, $\matr{Y}$, a 4D-tensor with dimensionality $N \times k \times u \times 1$ -- in both cases the last one refers to the single link travel time for each time-step/link combination.
 It is emphasized that each prediction consists of travel time predictions for all links for the next $k$ time-steps, i.e.\ multi-output, multi-time-step prediction.
@@
 The $N$ samples are sampled at a fixed time resolution, since we need a shared time reference across all links.
-\Cref{sec:experiments} elaborates on some of the considerations in choosing a sound resolution.
+\Cref{sec:experiments} elaborates on some of the considerations for choosing a sound resolution.
 
@@
   \centering
-  \includegraphics[scale=1.1]{drawings/data_shape.pdf}
+  \includegraphics[scale=1.0]{drawings/data_shape.pdf}
   \caption{Shapes of the input and output data.}
@@
 \subsection{Detrending}
-Urban bus travel times varies throughout the time of the day, and the day of the week due to \emph{recurring congestion}.
+Urban bus travel times vary throughout the time of the day, and the day of the week due to \emph{recurring congestion}.
 In order to reduce the need for the network to learn this recurring variation, link travel for link $\mathit{ln} \in \{ 1,\ldots,u\}$, at time-step $t$, $x_{\mathit{ln},t}$, is normalized to focus on deviations from the normal and expected pattern.
@@
 
-When calculating the mean and standard deviation it can be beneficial to exclude extreme outliers, since both mean and standard deviation are higly sensitive to such measurements.
+When calculating the mean and standard deviation it can be beneficial to exclude extreme outliers, since both mean and standard deviation are highly sensitive to such measurements.
 A suggested method is to apply \emph{absolute deviation around the median} (MAD) cf.~\cite{Olewuezi2011} when calculating $\mathit{\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}}$ and $\sigma_\mathit{ln}$.
@@
 
-During training the variables $\mathit{\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}}$ and $\sigma_\mathit{ln}$ should be calculated solely based on the training set, to emulated real-world application.
+During training the variables $\mathit{\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}}$ and $\sigma_\mathit{ln}$ should be calculated solely based on the training set, to emulate the real-world application.
+
+\section{Experiments}
+\label{sec:experiments}
+
+For the purpose of evaluation, the proposed method is applied to a dataset from Copenhagen's public transport authority: \emph{Movia}.
+The dataset consists of 1,2~M~travel time observations for the ``4A'' bus line in the period May to October 2017.
+The data points were collected using the real-time AVL-system installed in every vehicle servicing the line.
+
+\begin{figure}[!ht]
+  \centering
+  \includegraphics[width=0.4\textwidth]{drawings/map.png}
+  \caption{Geography of the 4A bus line in Copenhagen.}
+  \label{fig:4a_map}
+\end{figure}
+
+The geography of the route is shown in \Cref{fig:4a_map}.
+As the line circles Central Copenhagen, it is potentially highly sensitive to congestion to/from the city as it intersects with several large corridors along its route.
+Southeast of the city center the line splits in different destination patterns (gray), therefore only the first 32 links are considered in this experiment (red).
 
@@
 \end{figure*}
-\newpage
-
-\section{Experiments}
-\label{sec:experiments}
-
-For evaluation the method is applied to a dataset from Copenhagen's public transport authority, ``Movia''.
-The dataset consists of 1,2~M~travel time observations for the ``4A'' bus line in the period May to October 2017.
-The data points were collected using the real-time AVL-system installed in every vehicle servicing the line.
-
-\begin{figure}[!ht]
-  \centering
-  \includegraphics[width=0.4\textwidth]{drawings/map.png}
-  \caption{Geography of the 4A bus line in Copenhagen.}
-  \label{fig:4a_map}
-\end{figure}
-
-The geography of the route is shown in \Cref{fig:4a_map}.
-As the line circles Central Copenhagen, it is potentially highly sensitive to congestion to/from the city as it intersects with several large corridors along its route.
-Southeast of the city center the line splits in different destination patterns (gray), therefore only the first 32 links are considered in this experiment (red).
 
@@
 The black dots are actual measurements, and the lines the aggregated mean link travel time at the given resolution.
-The choice depends of the \emph{expected} frequency of the line, and is a balance between capturing the details and still having a reasonal number of measurements in each time-step to avoid over-fitting.
+The choice depends of the \emph{expected} frequency of the line, and is a balance between capturing the details and still having a reasonable number of measurements of each time-step to avoid over-fitting.
 
 For this experiment data was aggregated into 15-minute time-steps and normalized cf.~\Cref{sec:model}.
-This resolution were chosen based on a measured mean \emph{headway} of $7.5$ minutes between 06:00 and 22:00, i.e.\ the time between two vehicles during daytime.
+This resolution was chosen based on a measured mean \emph{headway} of $7.5$ minutes between 06:00 and 22:00, i.e.\ the time between two vehicles during daytime.
 
-%TODO: How much used for training, how much for testing?
+Given the time resolution, we set the fixed windows size, $w = 32$, equivalent of 8 hours, to allow pattern in the morning peek affect patterns in the afternoon peek.
+We set $k = 3$ to allow predictions of up to 45 minutes into the future.
+
 \subsection{Evaluation}
-In order to evaluate the presented model and comparisons, the following measures is used: \emph{mean absolute error} (MAE), \emph{root mean square error} (RMSE), and \emph{mean absolute percentage error} (MAPE) cf.~\cref{eq:mae,eq:rmse,eq:mape}, where $Y_i$ is the true link travel times for sample $i$ and $\widehat{Y}_i$ is the predicted travel time.
+In order to evaluate the proposed model and comparisons, the following measures are used: \emph{mean absolute error} (MAE), \emph{root mean square error} (RMSE), and \emph{mean absolute percentage error} (MAPE) cf.~\cref{eq:mae,eq:rmse,eq:mape}, where $\matr{Y}_i$ is the true link travel times for sample $i$ and $\matr{\widehat{Y}}_i$ is the predicted travel times.
+Since the multi-output, multi-time-step model predicts link travel times for all $u$ links for the next $k$ time steps, $\matr{Y}_i$ and $\matr{\widehat{Y}}_i$ have both the dimensionality $w \times u \times 1$.
 
 \begin{equation}
-    \textrm{MAE}(Y, \widehat{Y}) = \frac{\sum_{i = 1}^{N} \left| Y_i - \widehat{Y}_i \right| }{N}
+    \textrm{MAE}(\matr{Y}, \matr{\widehat{Y}}) = \frac{\sum_{i = 1}^{N} \left| \matr{Y}_i - \matr{\widehat{Y}}_i \right| }{N}
     \label{eq:mae}
@@
 \begin{equation}
-    \textrm{RMSE}(Y, \widehat{Y}) = \sqrt{\frac{\sum_{i = 1}^{N} \left(Y_i - \widehat{Y}_i \right)^2}{N}}
+    \textrm{RMSE}(\matr{Y}, \matr{\widehat{Y}}) = \sqrt{\frac{\sum_{i = 1}^{N} \left(\matr{Y}_i - \matr{\widehat{Y}}_i \right)^2}{N}}
     \label{eq:rmse}
@@
 \begin{equation}
-    \textrm{MAPE}(Y, \widehat{Y}) = \frac{1}{N} \sum_{i = 1}^{N} \left| \frac{Y_i - \widehat{Y}_i}{Y_i} \right| 
+    \textrm{MAPE}(\matr{Y}, \matr{\widehat{Y}}) = \frac{1}{N} \sum_{i = 1}^{N} \left| \frac{\matr{Y}_i - \matr{\widehat{Y}}_i}{\matr{Y}_i} \right| 
     \label{eq:mape} 
@@
     \end{tabular}
-    \caption{Results of the presented and other popular models}
+    \caption{Results of the proposed and the baseline models}
     \label{tab:results}
@@
 
-Since the multi-output, multi-time-step model predicts link travel times for all $u$ links for the next $k$ time steps, $Y_i$ and $\widehat{Y}_i$ are both $u \times k $ matrices.
-To allow a clear comparison, we reduce $Y_i$ and $\widehat{Y}_i$ by summing over all links cf.~\cref{eq:total}.
+To allow a clear comparison, we reduce $\matr{Y}_i$ and $\matr{\widehat{Y}}_i$ by summing over all links cf.~\cref{eq:total}.
 This is equivalent of predicting the total travel time of all 32 links, and follows the initial approach for arrival/departure time prediction by accumulating link travel times.
 \begin{align}
-    Y_i' = \sum_{\mathit{ln} = 1}^{u} Y_{i,\mathit{ln}} & & \widehat{Y}_i' = \sum_{\mathit{ln} = 1}^{u} \widehat{Y}_{i,\mathit{ln}}
+    \matr{Y}_i' = \sum_{\mathit{ln} = 1}^{u} \matr{Y}_{i,\mathit{ln}} & & \matr{\widehat{Y}}_i' = \sum_{\mathit{ln} = 1}^{u} \matr{\widehat{Y}}_{i,\mathit{ln}}
     \label{eq:total} 
 \end{align}
+
+The output of each of the evaluation functions is thus simply vector of size $k$, i.e.\ the evaluation of the different time steps for all links accumulated.
 
@@
 \label{sec:results}
-The performance of our \emph{ConvLSTM} model for link travel time prediction is compared several other models and services using the presented evaluation:
+The performance of our proposed model, based on \emph{ConvLSTM}, for link travel time prediction is compared against several other baseline models and services using the presented evaluation:
 \begin{enumerate}
    \item a naïve historical average model, i.e.\ equivalent of just predicting the normalized value, $\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}$ 
-   \item the current traffic prediction model deployed by ``Movia''.
+   \item the current traffic prediction model currently deployed by Movia.
    \item a pure LSTM-model for link travel time prediction, i.e.\ without applying convolutional filters in state transitions.
@@
 
-\begin{figure*}[!t]
-  \centering
-  \includegraphics[width=\textwidth]{plots/comparison_day.pdf}
-  \caption{}
-  \label{fig:comparison_day}
-\end{figure*}
+\Cref{tab:results} shows the overall performance of the proposed and the baseline models.
+Predictions are limited to daytime, i.e.\ between 06:00 and 22:00 and accumulated downstream on a journey level to simulate the use for real-time bus arrival/departure time prediction cf~\cref{eq:total}.
 
-\Cref{tab:results} shows the overall performance of the presented and the compared models.
-Predictions are limited to daytime, i.e.\ between 06:00 and 22:00 and accumulated downstream on journey level to simulate for use in real-time bus arrival/departure time prediction cf~\cref{eq:total}.
-
-Before going into a direct comparison, it is important to understand some aspects of the compared models, and how measurements were collected.
+Before going into a direct comparison, it is important to understand some aspects of the baseline models, and how measurements were collected.
 
 \subsection{Historical average}
-The performance of the historical average is independent with respect to the time steps ahead it predicts, as it just represents a weekly cycle of mean link travel times.
+The performance of the historical average is independent with respect to the number of time steps ahead it predicts, as it just represents a weekly cycle of mean link travel times.
 
 \subsection{Current model}
-Measurements for the currently deployed bus prediction model was collected at a 5-minute frequency using a non-publicly accessible endpoint at the transport authority.
+Measurements for the currently deployed bus prediction model were collected at a 5-minute frequency using a non-publicly accessible endpoint at the transport authority.
 The model is based on an historical average model, but has rule based mechanism on top, which can override or adjust the historical link travel times.
@@
 The model was trained on the exact same dataset as the \emph{ConvLSTM} model, and has a similar architecture, but without the convolutional filters.
-\newpage
 
@@
 Google uses crowd-sourced road congestion data collected from smart-phones with the \emph{Google Maps App} installed \citep{GMaps_Crowdsourcing}.
-While the exact model powering the service is not publicly described in details, the documentation states that ``estimates of travel time is given (with respect to) what is known about both historical traffic conditions and live traffic''.
-And further, ``that live traffic becomes more important the closer the departure time is to now''.
+While the exact model powering the service is not publicly described in detail, the documentation states that ``that the returned \emph{duration in traffic} should be the best estimate of travel time given what is known about both historical traffic conditions and live traffic''.
+And further, ``that live traffic becomes more important the closer the departure time is to now'' \citep{GMaps_DistanceMatrixAPI}.
 
-Because there is a limit on the number of request you can make freely to the API over a 24-hour period, it has only been possible to collect link travel times for the $t + 1$ time step (i.e.\ next 15 minutes).
+Because there is a limit on the number of requests that one freely can make to the API over a 24-hour period, it has only been possible to collect link travel times for the $t + 1$ time step (i.e.\ next 15 minutes).
+Travel times for each link were collected at a 15-minute interval between 06:00 and 22:00.
 
-Another important aspect is that the Google Traffic model is primary designed for estimating car driving times, and can thus be biased away from bus link travel times used in this experiment.
+Another important aspect is that the Google Traffic model is primary designed for estimating car travel times, and can thus be biased away from the bus travel times used in this experiment.
+Since we only consider link travel time, and collection data for each link individually, the bus dwell time will not be an issue, as it is not included in either measurement.
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=\textwidth]{plots/comparison_day.pdf}
+  \caption{}
+  \label{fig:comparison_day}
+\end{figure*}
+\begin{figure*}
+  \centering
+  \includegraphics[width=\textwidth]{plots/comparison_day_2.pdf}
+  \caption{}
+  \label{fig:comparison_day_2}
+\end{figure*}
 
 \subsection{Comparison}
-From the table it is found that the Convolutional LSTM performs better than the compared methods.
-Although the advantage might seem small it is emphasized that evaluation measurements are averaging their response, and thus the increased accuracy can be much higher on individual journeys, especially if they experienced very irregular travel times.
+We compare the performance of the proposed \emph{ConvLSTM} model for bus link travel time prediction against the presented baseline models.
+The overall results from \Cref{tab:results} shows that the \emph{ConvLSTM} model outperforms the other methods.
+The \emph{current model} performs the worst, even compared to the \emph{historical average} model, on which it is based.
+This is most likely due to the rule-based enforcement of recovery from delays, even though the data weights against a recovery.
 
-\subsection{Future work}
-For the prediction accuracy to be increased further it is proposed to apply ensemble/multi-model approaches.
-In this case the presented model can be included and used as a sub-model for the ensemble.
-Further research should be invested in the more rare, but highly impacting deviations, e.g.\ traffic incidents, extreme weather conditions, etc.
+Although the advantage might seem small it is emphasized that evaluation measurements are averaging their response, and thus the increased accuracy can be much higher on individual journeys, especially if they experience very irregular travel times.
+To investigate this, we focus our analysis on periods when the transport system is most vulnerable, and even small changes in regularity can propagate, since recovery is not an option, i.e.\ morning and afternoon peeks.
 
-\clearpage
+\begin{table}[!ht]
+  \center
+  \begin{tabular}{l|rrr}
+  Model &  RMSE &  MAE &  MAPE \\
+  \hline
+  \hline
+  Historical Average &        6.40 &       5.57 &     10.62 \% \\ \hline
+  Current Model      &        6.69 &       5.88 &     11.22 \% \\ \hline
+  Pure LSTM          &        0.00 &       0.00 &      0.00 \% \\ \hline
+  Google Traffic     &        5.25 &       4.62 &      9.17 \% \\ \hline
+  ConvLSTM           &        2.64 &       2.09 &      4.04 \% \\ \hline
+  \end{tabular}
+  \caption{Results: Morning peek (7--9)}
+  \label{tab:morning_peek}
+
+\end{table}
+
+\begin{table}[!ht]
+  \center
+  \begin{tabular}{l|rrr}
+  Model &  RMSE &  MAE &  MAPE \\
+  \hline
+  \hline
+  Historical Average &    5.90 &   4.65 &  8.28 \% \\ \hline
+  Current Model      &    6.28 &   5.20 &  9.37 \% \\ \hline
+  Pure LSTM          &    0.00 &   0.00 &  0.00 \% \\ \hline
+  Google Traffic     &    4.16 &   3.34 &  6.21 \% \\ \hline
+  ConvLSTM           &    3.79 &   3.02 &  5.61 \% \\ \hline
+  \end{tabular}
+  \caption{Results: Afternoon peek (14--18)}
+  \label{tab:afternoon_peek}  
+\end{table}
+
+\Cref{tab:morning_peek} and \Cref{tab:afternoon_peek} shows the evaluation results for \emph{morning peeks (weekdays, 7--9)} and \emph{afternoon peeks (weekdays, 14--18)} respectively for the time-step $t + 1$.
+
+The peek hour evaluation shows that the \emph{ConvLSTM} model increases its performance over the baseline models when the transport network is put under stress.
+In the morning peek, the \emph{ConvLSTM} model performs similar as the overall, whereas the baseline models all experience decreased performance of up to several minutes in both the RMSE and MAE, and an increased MAPE of roughly one third.
+
+Likewise the afternoon peek evaluation shows improvements with respect to the baseline models, even though the \emph{ConvLSTM} model also decreases its performance compared to the overall results.
+The outperformance is neither as significant as in the morning peek, and as the \emph{Google Traffic} model also performs quite well leaving the gap less than a minute.
+
+The get a little more detailed view of how the different models perform on the micro-level (i.e.\ the specific journey) we can inspect a single day of predictions.
+A random weekday from the test-dataset is plotted in~\Cref{fig:comparison_day}, which shows the accumulated travel time of all 32 links and the predicted travel time at time-step $t + 1$ both for the proposed and baseline models.
+
+In this particular day (a Thursday) the peek hour traffic was worse than normal, which leads both the \emph{historic averaging} model and the \emph{current model} to both underestimate travel time in the peek periods.
+Remember that the \emph{current model} is based on the \emph{historic averaging} model, so it is not unexpected that they perform very similar.
+There is also an small peek in travel time in the midday hours, which none of these historic averaging models are able to predict.
+
+On the other hand, both the \emph{Google Traffic} model and the proposed \emph{ConvLSTM} model get much closer to the ground truth in the peek ours.
+The \emph{Google Traffic} model seems to predict more accurate then \emph{ConvLSTM} model in the afternoon peek, whereas the opposite occurs in the morning peek.
+Even more interesting is the detection of the irregular peek in the midday hours, which both the \emph{Google Traffic} model and the proposed \emph{ConvLSTM} model are able to adjust to, at least to some degree.
+
+\Cref{fig:comparison_day_2}~shows another example (a Friday).
+Here the difference between the proposed model and the historical averaging and current model baselines are a bit weaker, simply because the the day to a larger degree follows the average pattern for a Friday (especially around the afternoon peek).
+Even though, the proposed model still performs the best, and it also supports the claim, that the proposed model is strongest when the traffic pattern deviates from the normal pattern, i.e.
+when the transport network is under stress.
 
@@
 \label{sec:conclusion}
-This paper has presented a multi-output, multi-time-step deep neural network for bus travel time prediction using Convolutional and Long short-term memory (LSTM) layers.
-Results shown the presented network outperforms other popular and recent methods.
+This paper has proposed a multi-output, multi-time-step deep neural network for bus travel time prediction using Convolutional and Long short-term memory (LSTM) layers.
+Results show, that the proposed model network outperforms other popular and recent methods.
+This includes Google's Traffic model based on crowd-sourced live data, and the current model deployed by Movia, the public transport authority in the Greater Copenhagen Area.
 
-Future research opportunities include using the model a part of a multi-model ensemble, and focus in the more rare, but highly impacting deviations.
+For the prediction accuracy to be increased further it is proposed to apply ensemble/multi-model approaches.
+In this case the proposed model can be included and used as a sub-model for the ensemble.
+Further research should be invested in the more rare, but highly impacting deviations, e.g.\ traffic incidents, extreme weather conditions, etc.
 
@@
 \bibliography{c:/development/references/library}
+
 \end{document}

### 2017-11-21 paper/paper.tex
Fix math notation.
Final draft

@@
     \center
-    Department of Management Engineering \\
-    Technical University of Denmark, 2800 Kongens Lyngby 
+    Public Transport Movia \\
+    Gammel Køge Landevej 3, DK-2500 Valby
   \end{minipage}% This must go next to `\end{minipage}`
@@
     \center
-    Public Transport Movia \\
-    Gammel Køge Landevej 3, 2500 Valby 
+    Department of Management Engineering \\
+    Technical University of Denmark, DK-2800 Kongens Lyngby
   \end{minipage}
@@
 This paper presents a multi-output, multi-time-step, deep neural network for bus travel time prediction using Convolutional and Long short-term memory (LSTM) layers.
-The method is empirically evaluated and compared to other popular approaches for link travel time prediction, and currently available services.
-We find that the proposed model significantly outperforms all the other methods, and is able to detect small irregular peeks in bus travel time very quickly.
+The method is empirically evaluated and compared to other popular approaches for link travel time prediction, and currently available services, including the currently deployed model.
+We find that the proposed model significantly outperforms all the other methods, and is able to detect small irregular peaks in bus travel time very quickly.
 \end{abstract}
@@
 The method is empirically evaluated and compared to other popular approaches for link travel time prediction, including the currently model deployed in production by the public transport authority for the Greater Copenhagen Area: \emph{Movia}.
-Furthermore the method is compared to Google Traffic, a popular online service for travel time prediction.
+Furthermore the method is compared to Google Traffic (part of Google Maps), a popular online service for travel time prediction.
 
@@
 This is substantiated by \cite{Lin2013} and \cite{Kumar2014}, who find artificial neural networks (ANN) to outperform Kalman-filter models.
-However, the computational challenges of fully connected ANNs is also limiting the depth of the network and thus how complex patterns it can learn to recognize.
+However, the computational challenges of fully connected ANNs are also limiting the depth of the network and thus how complex patterns it can learn to recognize.
 This has sparked the interest for studying composite or hybrid models.
-\citet{Bai2015} uses a two-stage approach by combining offline ANN-models, with an adaptable/online Kalman-filter to yield a dynamic model.
+\citet{Bai2015} uses a two-stage approach by combining an offline ANN-model, with an adaptable/online Kalman-filter to yield a dynamic model.
 The model is not actually learning in the long term, but is able to adapt to temporal variations in the current travel time on a journey, and is still not able to recognize long distance patterns.
@@
 To allow this mechanism the persistence of information is controlled by three gates: \emph{input gate}, \emph{forget gate}, and \emph{output gate}.
-Each gate yields a state value at time $t$, respectively $\matr{i}_t$, $\matr{f}_t$, and $\matr{o}_t$, along with the cell output, $\matr{h}_t$, cf.~\cref{eq:lstm}, where $\circ$ denotes the entry-wise product.
+Each gate yields a state variable at time $t$, respectively $\matr{i}_t$, $\matr{f}_t$, and $\matr{o}_t$, along with the cell output, $\matr{h}_t$, cf.~\cref{eq:lstm}, where $\circ$ denotes the entry-wise product.
 
@@
   \matr{f}_t &= \sigma \left( \matr{W}^f \matr{x}_t + \matr{R}^f \matr{h}_{t-1} + \matr{U}^f \circ \matr{c}_{t-1} + \matr{b}^f \right) \\
-  \matr{c}_t &= \matr{f}_t \circ \matr{c}_{t-1} + \matr{i}_t \circ \mathrm{tanh} \left( \matr{W}^c \matr{x}_t + \matr{R}^c \matr{h}_{t-1} + \matr{b}_c \right) \\
+  \matr{c}_t &= \matr{f}_t \circ \matr{c}_{t-1} + \matr{i}_t \circ \mathrm{tanh} \left( \matr{W}^c \matr{x}_t + \matr{R}^c \matr{h}_{t-1} + \matr{b}^c \right) \\
   \matr{o}_t &= \sigma \left( \matr{W}^\mathit{o} \matr{x}_t + \matr{R}^\mathit{o} \matr{h}_{t-1} + \matr{U}^\mathit{o} \circ \matr{c}_{t} + \matr{b}^o \right) \\
@@
 
-Convolutional Neural Networks~(CNN) on the other hand have been widely used for capturing spacial relationships, e.g.\ importance of neighboring pixels in an image.
+Convolutional Neural Networks~(CNN) on the other hand have been widely used for capturing spatial relationships, e.g.\ importance of neighboring pixels in an image.
 As opposed to fully connected layers, where each unit, $i$, in the layer, has a dedicated scalar weight, $w_{ij}$, for all input value, $x_j$, convolutional units are only locally connected and reuse the same weights for producing several outputs.
@@
   \matr{f}_t &= \sigma \left( \matr{W}^f * \matr{x}_t + \matr{R}^f * \matr{h}_{t-1} + \matr{U}^f \circ \matr{c}_{t-1} + \matr{b}^f \right) \\
-  \matr{c}_t &= \matr{f}_t \circ \matr{c}_{t-1} + \matr{i}_t \circ \mathrm{tanh} \left( \matr{W}^c * \matr{x}_t + \matr{R}^c * \matr{h}_{t-1} + \matr{b}_c \right) \\
+  \matr{c}_t &= \matr{f}_t \circ \matr{c}_{t-1} + \matr{i}_t \circ \mathrm{tanh} \left( \matr{W}^c * \matr{x}_t + \matr{R}^c * \matr{h}_{t-1} + \matr{b}^c \right) \\
   \matr{o}_t &= \sigma \left( \matr{W}^\mathit{o} * \matr{x}_t + \matr{R}^\mathit{o} * \matr{h}_{t-1} + \matr{U}^\mathit{o} \circ \matr{c}_{t} + \matr{b}^o \right) \\
@@
 This size is used both for the \emph{input-to-state} and \emph{state-to-state} convolutional filters.
-Finally each \emph{ConvLSTM} layer 64~outputs, yielding a total of 512 convolutional filters.
+Finally each \emph{ConvLSTM} layer has 64~outputs, yielding a total need of 512 convolutional filters.
 
@@
 
-For the \emph{ConvLSTM} model to be able to capture the desired spatial-–temporal patterns, the input data must be arranged in a suitable manner: I.e.\ in $N$ samples, each with a window of the $w$ lagging time-steps $t-w+1, \ldots, t$, and each time-step with $u$ link travel times $1, \ldots, u$ cf.~\Cref{fig:data_shape}.
+For the \emph{ConvLSTM} model to be able to capture the desired spatial--temporal patterns, the input data must be arranged in a suitable manner: I.e.\ in $N$ samples, each with a window of the $w$ lagging time-steps $t-w+1, \ldots, t$, and each time-step with $u$ link travel times $1, \ldots, u$ cf.~\Cref{fig:data_shape}.
 
@@
 Urban bus travel times vary throughout the time of the day, and the day of the week due to \emph{recurring congestion}.
-In order to reduce the need for the network to learn this recurring variation, link travel for link $\mathit{ln} \in \{ 1,\ldots,u\}$, at time-step $t$, $x_{\mathit{ln},t}$, is normalized to focus on deviations from the normal and expected pattern.
+In order to reduce the need for the network to learn this recurring variation, link travel time for link $\mathit{ln} \in \{ 1,\ldots,u\}$, at time-step $t$, $x_{\mathit{ln},t}$, is normalized to focus on deviations from the normal and expected pattern.
 Travel times are centered with the mean for each link, at the time of day, and day of week, $\mathit{\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}}$, and scaled with the standard deviation for each link, $\sigma_\mathit{ln}$, cf.~\cref{eq:normalization}.
@@
 
-Given the time resolution, we set the fixed windows size, $w = 32$, equivalent of 8 hours, to allow pattern in the morning peek affect patterns in the afternoon peek.
+Given the time resolution, we set the fixed window size, $w = 32$, equivalent of 8 hours, to allow pattern in the morning peak affect patterns in the afternoon peak.
 We set $k = 3$ to allow predictions of up to 45 minutes into the future.
@@
    \item a naïve historical average model, i.e.\ equivalent of just predicting the normalized value, $\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}$ 
-   \item the current traffic prediction model currently deployed by Movia.
+   \item the traffic prediction model currently deployed by Movia.
    \item a pure LSTM-model for link travel time prediction, i.e.\ without applying convolutional filters in state transitions.
@@
 Measurements for the currently deployed bus prediction model were collected at a 5-minute frequency using a non-publicly accessible endpoint at the transport authority.
-The model is based on an historical average model, but has rule based mechanism on top, which can override or adjust the historical link travel times.
+The model is based on an historical average model, but also has a rule based mechanism on top, which can override or adjust the historical link travel times.
 For instance it will assume that a delayed vehicle can recover (partially or completely) from its delay by traversing links faster.
@@
 Google uses crowd-sourced road congestion data collected from smart-phones with the \emph{Google Maps App} installed \citep{GMaps_Crowdsourcing}.
-While the exact model powering the service is not publicly described in detail, the documentation states that ``that the returned \emph{duration in traffic} should be the best estimate of travel time given what is known about both historical traffic conditions and live traffic''.
+While the exact model powering the service is not publicly described in detail, the documentation states that ``the returned \emph{duration in traffic} should be the best estimate of travel time given what is known about both historical traffic conditions and live traffic''.
 And further, ``that live traffic becomes more important the closer the departure time is to now'' \citep{GMaps_DistanceMatrixAPI}.
@@
 Another important aspect is that the Google Traffic model is primary designed for estimating car travel times, and can thus be biased away from the bus travel times used in this experiment.
-Since we only consider link travel time, and collection data for each link individually, the bus dwell time will not be an issue, as it is not included in either measurement.
+Since we only consider link travel time, and collects data for each link individually, the bus dwell time will not be an issue, as it is not included in either measurement.
 
@@
 Although the advantage might seem small it is emphasized that evaluation measurements are averaging their response, and thus the increased accuracy can be much higher on individual journeys, especially if they experience very irregular travel times.
-To investigate this, we focus our analysis on periods when the transport system is most vulnerable, and even small changes in regularity can propagate, since recovery is not an option, i.e.\ morning and afternoon peeks.
+To investigate this, we focus our analysis on periods when the transport system is most vulnerable, and even small changes in regularity can propagate, since recovery is not an option, i.e.\ morning and afternoon peaks.
 
@@
   \end{tabular}
-  \caption{Results: Morning peek (7--9)}
-  \label{tab:morning_peek}
+  \caption{Results: Morning peak (7--9)}
+  \label{tab:morning_peak}
 
@@
   \end{tabular}
-  \caption{Results: Afternoon peek (14--18)}
-  \label{tab:afternoon_peek}  
+  \caption{Results: Afternoon peak (14--18)}
+  \label{tab:afternoon_peak}  
 \end{table}
 
-\Cref{tab:morning_peek} and \Cref{tab:afternoon_peek} shows the evaluation results for \emph{morning peeks (weekdays, 7--9)} and \emph{afternoon peeks (weekdays, 14--18)} respectively for the time-step $t + 1$.
+\Cref{tab:morning_peak} and \Cref{tab:afternoon_peak} shows the evaluation results for \emph{morning peaks (weekdays, 7--9)} and \emph{afternoon peaks (weekdays, 14--18)} respectively for the time-step $t + 1$.
 
-The peek hour evaluation shows that the \emph{ConvLSTM} model increases its performance over the baseline models when the transport network is put under stress.
-In the morning peek, the \emph{ConvLSTM} model performs similar as the overall, whereas the baseline models all experience decreased performance of up to several minutes in both the RMSE and MAE, and an increased MAPE of roughly one third.
+The peak hour evaluation shows that the \emph{ConvLSTM} model increases its performance over the baseline models when the transport network is put under stress.
+In the morning peak, the \emph{ConvLSTM} model performs similar as the overall, whereas the baseline models all experience decreased performance of up to several minutes in both the RMSE and MAE, and an increased MAPE of roughly one third.
 
-Likewise the afternoon peek evaluation shows improvements with respect to the baseline models, even though the \emph{ConvLSTM} model also decreases its performance compared to the overall results.
-The outperformance is neither as significant as in the morning peek, and as the \emph{Google Traffic} model also performs quite well leaving the gap less than a minute.
+Likewise the afternoon peak evaluation shows improvements with respect to the baseline models, even though the \emph{ConvLSTM} model also decreases its performance compared to the overall results.
+The outperformance is neither as significant as in the morning peak, and as the \emph{Google Traffic} model also performs quite well leaving the gap less than a minute.
 
-The get a little more detailed view of how the different models perform on the micro-level (i.e.\ the specific journey) we can inspect a single day of predictions.
+To get a more detailed view of how the different models perform on the micro-level (i.e.\ the specific journey) we can inspect a single day of predictions.
 A random weekday from the test-dataset is plotted in~\Cref{fig:comparison_day}, which shows the accumulated travel time of all 32 links and the predicted travel time at time-step $t + 1$ both for the proposed and baseline models.
 
-In this particular day (a Thursday) the peek hour traffic was worse than normal, which leads both the \emph{historic averaging} model and the \emph{current model} to both underestimate travel time in the peek periods.
+In this particular day (a Thursday) the peak hour traffic was worse than normal, which leads both the \emph{historic averaging} model and the \emph{current model} to both underestimate travel time in the peak periods.
 Remember that the \emph{current model} is based on the \emph{historic averaging} model, so it is not unexpected that they perform very similar.
-There is also an small peek in travel time in the midday hours, which none of these historic averaging models are able to predict.
+There is also small peak in travel time in the midday hours, which none of these historic averaging models are able to predict.
 
-On the other hand, both the \emph{Google Traffic} model and the proposed \emph{ConvLSTM} model get much closer to the ground truth in the peek ours.
-The \emph{Google Traffic} model seems to predict more accurate then \emph{ConvLSTM} model in the afternoon peek, whereas the opposite occurs in the morning peek.
-Even more interesting is the detection of the irregular peek in the midday hours, which both the \emph{Google Traffic} model and the proposed \emph{ConvLSTM} model are able to adjust to, at least to some degree.
+On the other hand, both the \emph{Google Traffic} model and the proposed \emph{ConvLSTM} model get much closer to the ground truth in the peak hours.
+The \emph{Google Traffic} model seems to predict more accurate then \emph{ConvLSTM} model in the afternoon peak, whereas the opposite occurs in the morning peak.
+Even more interesting is the detection of the irregular peak in the midday hours, which both the \emph{Google Traffic} model and the proposed \emph{ConvLSTM} model are able to adjust to, at least to some degree.
 
 \Cref{fig:comparison_day_2}~shows another example (a Friday).
-Here the difference between the proposed model and the historical averaging and current model baselines are a bit weaker, simply because the the day to a larger degree follows the average pattern for a Friday (especially around the afternoon peek).
-Even though, the proposed model still performs the best, and it also supports the claim, that the proposed model is strongest when the traffic pattern deviates from the normal pattern, i.e.
-when the transport network is under stress.
+Here the difference between the proposed model and the historical averaging and current model baselines are a bit weaker, simply because the the day to a larger degree follows the average pattern for a Friday (especially around the afternoon peak).
+Even though, the proposed model still performs the best, and this also supports the claim, that the proposed model is strongest when the traffic pattern deviates from the normal pattern, i.e.\ when the transport network is under stress.
 

### 2017-11-26 paper/paper.tex
Draft2

@@
 \usepackage{subcaption}
-\usepackage{lineno}
 
@@
 %% \author[label1,label2]{}
-%% \address[label1]{}
-%% \address[label2]{}
+\address[movia]{Public Transport Movia, Gammel Køge Landevej 3, 2500 Valby, Denmark}
+\address[dtu]{Department of Management Engineering, Technical University of Denmark, 2800 Kongens Lyngby, Denmark}
 
-\author{Niklas Christoffer Petersen\\[.25em]\footnotesize{niklch@dtu.dk}}
+\author[movia,dtu]{Niklas Christoffer Petersen\corref{cor1}}
+\ead{niklch@dtu.dk}
+\author[dtu]{Filipe Rodrigues}
+\ead{rodr@dtu.dk}
+\author[dtu]{Francisco Camara Pereira}
+\ead{camara@dtu.dk}
+\cortext[cor1]{Corresponding author}
 
-\address{
-  \vspace{.5em}
-  \begin{minipage}{.5\textwidth}
-    \center
-    Public Transport Movia \\
-    Gammel Køge Landevej 3, DK-2500 Valby
-  \end{minipage}% This must go next to `\end{minipage}`
-  \begin{minipage}{.5\textwidth}
-    \center
-    Department of Management Engineering \\
-    Technical University of Denmark, DK-2800 Kongens Lyngby
-  \end{minipage}
-  \vspace{-2.5em}
-}
+%\\[.25em]\footnotesize{niklch}
 
 \begin{abstract}
-Accurate and reliable predictions for travel times in public transport networks is essential for delivering an attractive service.
+Accurate and reliable predictions for travel times in public transport networks are essential for delivering an attractive service.
 This paper presents a multi-output, multi-time-step, deep neural network for bus travel time prediction using Convolutional and Long short-term memory (LSTM) layers.
-The method is empirically evaluated and compared to other popular approaches for link travel time prediction, and currently available services, including the currently deployed model.
+The method is empirically evaluated and compared to other popular approaches for link travel time prediction, and currently available services, including the currently deployed model at Movia, the regional public transport authority in Greater Copenhagen.
 We find that the proposed model significantly outperforms all the other methods, and is able to detect small irregular peaks in bus travel time very quickly.
@@
 %% keywords here, in the form: keyword \sep keyword
-Bus Travel Time Prediction \sep Intelligent Transport Systems \sep Convolutional Neural Network (CNN) \sep Long short-term memory (LSTM), Deep Learning.
+Bus Travel Time Prediction \sep Intelligent Transport Systems \sep Convolutional Neural Network (CNN) \sep Long short-term memory (LSTM) \sep Deep Learning.
 \end{keyword}
@@
 Examples include real-time traffic information for passengers, e.g.\ smartphone travel planner apps, in-vehicle infotainment screens and departure boards.
-Studies has shown that reliable real-time information at bus stops has a statistical significant dampening effect on the perceived waiting time \cite{Fan2016}.
+Studies have shown that reliable real-time information at bus stops has a statistical significant dampening effect on the perceived waiting time \cite{Fan2016}.
 
@@
 The predicted travel time for each link is simply accumulated downstream the route to yield the arrival/departure time predictions at each stop point of the rest of the current journey.
-Besides the link travel time, estimations of dwell time (i.e.\ when a bus is holding at a stop point) are also accumulated downstream.
+Thus, in the example, to predict when the bus at stop A will arrive at stop B, we would just sum our predicted link travel times for Link 1 to 3.
+Besides the link travel time, estimations of dwell time (i.e.\ when a bus is holding at a stop point) should also be accumulated downstream.
 
@@
 The problem becomes much more complex in urban areas where congestion, events, road-works, weather, etc.\ highly influence the traffic flow and passenger demand.
-As on-board GPS and AVL-systems has become more affordable and common, data has both grown in coverage, i.e.\ number of vehicles with AVL installed, and frequency, i.e.\ number of GPS-positions collected per vehicle.
+As on-board GPS and AVL-systems have become more affordable and common, data has both grown in coverage, i.e.\ number of vehicles with AVL installed, and frequency, i.e.\ number of GPS-positions collected per vehicle.
 
 In this paper we present a multi-output, multi-time-step deep neural network for bus travel time prediction using a combination of \emph{Convolutional} and \emph{Long short-term memory} (LSTM) \cite{Lstm1,Lstm2} layers.
-The goal of this work is to produce precise short-term predictions (e.g.\ $0-1.5$ hours) for link travel time, specifically for bus traffic in urban areas.
+The goal of this work is to produce precise short-term predictions (e.g.
+up to\ $0-1.5$ hours in the future) for link travel time, specifically for bus traffic in urban areas.
 The method is empirically evaluated and compared to other popular approaches for link travel time prediction, including the currently model deployed in production by the public transport authority for the Greater Copenhagen Area: \emph{Movia}.
@@
 Bus link travel time prediction has been explored in research as GPS and AVL data has become increasingly available.
-The problem overlaps with other research areas such as general traffic flow and speed estimation, but also has unique constraints and opportunities that follows from servicing a fixed route.
+The problem overlaps with other research areas such as general traffic flow and speed estimation, but also has unique constraints and opportunities that follow from servicing a fixed route.
 The improvement in computational power in the recent decades has gradually allowed more complex link travel time models with increased precision.
@@
 Early approaches for bus travel time prediction rely on historical averaging models \cite{Dailey1999,Sun2007}, and linear regression \cite{Patnaik2004}.
-Recent research presents this type of model only for comparison purposes, and the models are in all cases outperformed by the proposed models \cite{Shalaby2004,Jeong2005}.
+Recent research presents this type of models only for comparison purposes, and in all cases, these are outperformed by the proposed alternatives \cite{Shalaby2004,Jeong2005}.
 A historical averaging model will only slowly converge to changes in the travel time, which of cause is undesired with short, but highly impacting, external influences (e.g.\ a traffic incident or large event).
 
-Kalman-filters, which by its capabilities of maintaining state between predictions, has shown interest from several studies for capturing \emph{changes} in the travel time, either as an independent model \cite{Chen2001,Shalaby2004} or in combination with other models \cite{Yo2010,Bai2015}.
+Kalman-filters, which by their capabilities of maintaining state between predictions, have shown interest from several studies for capturing \emph{changes} in the travel time, either as an independent model \cite{Chen2001,Shalaby2004} or in combination with other models \cite{Yo2010,Bai2015}.
 The Kalman-filter models still have limited options for capturing and forecasting the complex dynamics of travel and dwell time in a metropolitan bus system.
@@
 In contrast the combination of both \emph{LSTM} cells and \emph{Convolutional} filters for bus travel time prediction, proposed in this paper, allows the learned patterns to generalize beyond a single link and time, i.e.\ multi-output and multi-time-step.
-Furthermore this reduces the computational complexity by magnitudes compared to fully connected ANNs capable of capturing similar complex patterns.
+Furthermore this reduces the computational complexity by orders of maginitude compared to fully connected ANNs capable of capturing similar complex patterns.
 
@@
 To allow this mechanism the persistence of information is controlled by three gates: \emph{input gate}, \emph{forget gate}, and \emph{output gate}.
-Each gate yields a state variable at time $t$, respectively $\matr{i}_t$, $\matr{f}_t$, and $\matr{o}_t$, along with the cell output, $\matr{h}_t$, cf.~\cref{eq:lstm}, where $\circ$ denotes the entry-wise product.
+Each gate yields a state variable at time $t$, respectively $\matr{i}_t$, $\matr{f}_t$, and $\matr{o}_t$, along with the cell output, $\matr{h}_t$, cf.~\cref{eq:lstm}, where $\circ$ denotes the element-wise product.
 
@@
   \centering
-  \includegraphics[scale=.45]{drawings/conv.pdf}
+  \includegraphics[scale=.5]{drawings/conv.pdf}
   \caption{Application of Convolutional filter onto 1D data.}
@@
 Special care needs to be taken on the boundaries, i.e.\ where the convolutional filter will exceed the input.
-A popular approach to avoid the size of the output decreases is to pad the input, e.g.\ with zeros.
+To avoid that the size of the output decreases, an approach is to pad the input, e.g.\ with zeros.
 This ensures that the output shape of each convolutional unit will always be identical to the input shape, which is often desirable.
@@
 This allows for even deeper networks.
-\newpage
 
@@
 
-Given the time resolution, we set the fixed window size, $w = 32$, equivalent of 8 hours, to allow pattern in the morning peak affect patterns in the afternoon peak.
+Given the time resolution, we set the fixed window size, $w = 32$, equivalent of 8 hours, to allow patterns in the morning peak affect patterns in the afternoon peak.
 We set $k = 3$ to allow predictions of up to 45 minutes into the future.
@@
 Measurements for the currently deployed bus prediction model were collected at a 5-minute frequency using a non-publicly accessible endpoint at the transport authority.
-The model is based on an historical average model, but also has a rule based mechanism on top, which can override or adjust the historical link travel times.
+The model is based on a historical average model, but also has a rule based mechanism on top, which can override or adjust the historical link travel times.
 For instance it will assume that a delayed vehicle can recover (partially or completely) from its delay by traversing links faster.
@@
   \includegraphics[width=\textwidth]{plots/comparison_day.pdf}
-  \caption{}
+  \caption{Accumulated link travel Time over a single day (a Thursday) in the test set.}
   \label{fig:comparison_day}
@@
   \includegraphics[width=\textwidth]{plots/comparison_day_2.pdf}
-  \caption{}
+  \caption{Accumulated link travel Time over a single day (a Friday) in the test set.}
   \label{fig:comparison_day_2}
@@
 The \emph{current model} performs the worst, even compared to the \emph{historical average} model, on which it is based.
-This is most likely due to the rule-based enforcement of recovery from delays, even though the data weights against a recovery.
+This is most likely due to the rule-based enforcement of recovery from delays, even though the data weighs against a recovery.
 
@@
   Current Model      &        6.69 &       5.88 &     11.22 \% \\ \hline
-  Pure LSTM          &        0.00 &       0.00 &      0.00 \% \\ \hline
+  Pure LSTM          &        3.80 &       3.16 &      6.01 \% \\ \hline
   Google Traffic     &        5.25 &       4.62 &      9.17 \% \\ \hline
@@
   Current Model      &    6.28 &   5.20 &  9.37 \% \\ \hline
-  Pure LSTM          &    0.00 &   0.00 &  0.00 \% \\ \hline
+  Pure LSTM          &    5.26 &       3.97 &      7.08 \% \\ \hline
   Google Traffic     &    4.16 &   3.34 &  6.21 \% \\ \hline
@@
 
-\Cref{tab:morning_peak} and \Cref{tab:afternoon_peak} shows the evaluation results for \emph{morning peaks (weekdays, 7--9)} and \emph{afternoon peaks (weekdays, 14--18)} respectively for the time-step $t + 1$.
+\Cref{tab:morning_peak} and \Cref{tab:afternoon_peak} show the evaluation results for \emph{morning peaks (weekdays, 7--9)} and \emph{afternoon peaks (weekdays, 14--18)} respectively for the time-step $t + 1$.
 
 The peak hour evaluation shows that the \emph{ConvLSTM} model increases its performance over the baseline models when the transport network is put under stress.
-In the morning peak, the \emph{ConvLSTM} model performs similar as the overall, whereas the baseline models all experience decreased performance of up to several minutes in both the RMSE and MAE, and an increased MAPE of roughly one third.
+In the morning peak, the \emph{ConvLSTM} model does not degrade in performormance compared to the overall daytime results, whereas the baseline models all experience decreased performance of up to several minutes in both the RMSE and MAE, and an increased MAPE of roughly one third.
 
@@
 Remember that the \emph{current model} is based on the \emph{historic averaging} model, so it is not unexpected that they perform very similar.
-There is also small peak in travel time in the midday hours, which none of these historic averaging models are able to predict.
+There is also a small peak in travel time in the midday hours, which none of these historic averaging models are able to predict.
 

### 2017-11-27 map/convert.js
Fix path to map.html.

@@ -4,7 +4,7 @@ const puppeteer = require('puppeteer');
   const browser = await puppeteer.launch();
   const page = await browser.newPage();
   await page.setViewport({width: 800, height: 1000});
-  await page.goto('file://C:/Users/ncp/Jupyter/bus-arrival-convlstm/map/map.html');
+  await page.goto('file://./map.html');
   await page.screenshot({path: 'map.png'});
 
   await browser.close();
